---
title       : Using data.table
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=FALSE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## data.table

* Inherets from data.frame
  * All functions that accept data.frame work on data.table
* Written in C so it is much faster
* Much, much faster at subsetting, group, and updating
 
---

## Create data tables just like data frames

```{r init}

library(data.table)
DF = data.frame(x=rnorm(9),y=rep(c("a","b","c"),each=3),z=rnorm(9))
DT = data.table(x=rnorm(9),y=rep(c("a","b","c"),each=3),z=rnorm(9)); DT
             x y           z
1:  0.56884686 a  0.12221856
2:  1.53922467 a  1.94066509
3:  1.34163075 a -0.89852855
4: -0.22804722 b  0.04096248
5: -1.87879805 b  0.72378938
6: -0.03807838 b  0.40187776
7: -1.84764149 c -0.55463610
8: -0.71876726 c -0.33501729
9:  1.82553854 c -0.05447002

head(DT,3)
           x y          z
1 -1.6721944 a -0.4627480
2  0.2040072 a  0.6223860
3 -0.3785383 a  0.2554901
4  0.2736031 b -1.2430462
5 -1.4866092 b -0.2462717
6 -1.8493293 b  1.7116482
7  1.1078240 c  0.1096962
8 -0.1259785 c  3.2043162
9  0.6416656 c -0.1443558

```

## See all the data tables in memory

```{r,dependson="init"}

tables()
     NAME NROW MB COLS  KEY
[1,] DT      9 1  x,y,z    
Total: 1MB

```

## Subsetting rows

```{r,Jeff=Getting data}

DT[2,]
          x y        z
1: 1.539225 a 1.940665

DT[DF$y=="a",]
            x y          z
1: 0.5688469 a  0.1222186
2: 1.5392247 a  1.9406651
3: 1.3416307 a -0.8985285

```

## Subsetting rows

```{r,dependson="init"}

DT[c(2,3)]
          x y          z
1: 1.539225 a  1.9406651
2: 1.341631 a -0.8985285

```

## Subsetting columns!?

```{r,dependson="init"}

DT[,c(2,3)]
  y          z
          x y          z
1: 1.539225 a  1.9406651
2: 1.341631 a -0.8985285
> DT[,c(2,3)]
[1] 2 3

```

## Column subsetting in data.table

* The subsetting function is modified for data.table
* The argument you pass after the comma is called an "expression"
* In R an expression is a collection of statements enclosed in curley brackets 

```{r }

{
  x = 1
  y = 2
}
k = {print(10); 5}
[1] 10
print(k)
[1] 5

```

## Calculating values for variables with expressions

```{r,dependson="init"}

DT[,list(mean(x),sum(z))]
          V1       V2
1: 0.06265649 1.386861

DT[,table(y)]
y
a b c 
3 3 3 
```

## Adding new columns

```{r,dependson="init"}

DT[,w:=z^2]; DT
             x y           z           w
1:  0.56884686 a  0.12221856 0.014937376
2:  1.53922467 a  1.94066509 3.766181002
3:  1.34163075 a -0.89852855 0.807353547
4: -0.22804722 b  0.04096248 0.001677925
5: -1.87879805 b  0.72378938 0.523871072
6: -0.03807838 b  0.40187776 0.161505734
7: -1.84764149 c -0.55463610 0.307621199
8: -0.71876726 c -0.33501729 0.112236586
9:  1.82553854 c -0.05447002 0.002966983

```

## Adding new columns

```{r dt2,dependson="init"}

DT2 <- DT
DT[, y:= 2]
   # Warning message

```


## Careful

```{r ,dependson="dt2"}

head(DT,n=3)
           x y          z          w
1: 0.5688469 2  0.1222186 0.01493738
2: 1.5392247 2  1.9406651 3.76618100
3: 1.3416307 2 -0.8985285 0.80735355

head(DT2,n=3)
           x y          z          w
1: 0.5688469 2  0.1222186 0.01493738
2: 1.5392247 2  1.9406651 3.76618100
3: 1.3416307 2 -0.8985285 0.80735355

```

## Multiple operations

```{r,dependson="init"}

DT[,m:= {tmp <- (x+z); log2(tmp+5)}]; DT
             x y           z           w        m
1:  0.56884686 2  0.12221856 0.014937376 2.508699
2:  1.53922467 2  1.94066509 3.766181002 3.084046
3:  1.34163075 2 -0.89852855 0.807353547 2.444429
4: -0.22804722 2  0.04096248 0.001677925 2.266911
5: -1.87879805 2  0.72378938 0.523871072 1.942980
6: -0.03807838 2  0.40187776 0.161505734 2.423255
7: -1.84764149 2 -0.55463610 0.307621199 1.377247
8: -0.71876726 2 -0.33501729 0.112236586 1.980470
9:  1.82553854 2 -0.05447002 0.002966983 2.759384

```

---

## plyr like operations

```{r,dependson="init"}

DT[,a:=x>0]
             x y           z           w        m     a
1:  0.56884686 2  0.12221856 0.014937376 2.508699  TRUE
2:  1.53922467 2  1.94066509 3.766181002 3.084046  TRUE
3:  1.34163075 2 -0.89852855 0.807353547 2.444429  TRUE
4: -0.22804722 2  0.04096248 0.001677925 2.266911 FALSE
5: -1.87879805 2  0.72378938 0.523871072 1.942980 FALSE
6: -0.03807838 2  0.40187776 0.161505734 2.423255 FALSE
7: -1.84764149 2 -0.55463610 0.307621199 1.377247 FALSE
8: -0.71876726 2 -0.33501729 0.112236586 1.980470 FALSE
9:  1.82553854 2 -0.05447002 0.002966983 2.759384  TRUE
```

---

## plyr like operations

```{r,dependson="init"}

DT[,b:= mean(x+w),by=a]
            x y           z           w        m     a         b
1:  0.56884686 2  0.12221856 0.014937376 2.508699  TRUE  2.466670
2:  1.53922467 2  1.94066509 3.766181002 3.084046  TRUE  2.466670
3:  1.34163075 2 -0.89852855 0.807353547 2.444429  TRUE  2.466670
4: -0.22804722 2  0.04096248 0.001677925 2.266911 FALSE -0.720884
5: -1.87879805 2  0.72378938 0.523871072 1.942980 FALSE -0.720884
6: -0.03807838 2  0.40187776 0.161505734 2.423255 FALSE -0.720884
7: -1.84764149 2 -0.55463610 0.307621199 1.377247 FALSE -0.720884
8: -0.71876726 2 -0.33501729 0.112236586 1.980470 FALSE -0.720884
9:  1.82553854 2 -0.05447002 0.002966983 2.759384  TRUE  2.466670

```

---

## Special variables

`.N` An integer, length 1, containing the number of elements of a factor level 

```{r}

set.seed(123);
DT <- data.table(x=sample(letters[1:3], 1E5, TRUE))
DT[, .N, by=x]
   x     N
1: a 33387
2: c 33201
3: b 33412

```

---

## Keys

```{r}
DT <- data.table(x=rep(c("a","b","c"),each=100), y=rnorm(300)); DT
     x          y
  1: a  0.2595897
  2: a  0.9175107
  3: a -0.7223183
  4: a -0.8082840
  5: a -0.1413520
.......             
296: c  1.6064561
297: c  1.1905645
298: c -2.1388180
299: c  0.5035984
300: c -0.4108657
setkey(DT, x)
DT['a']
    x           y
  1: a  0.25958973
  2: a  0.91751072
  3: a -0.72231834
  4: a -0.80828402
  5: a -0.14135202
  6: a  2.25701345
  7: a -2.37955015
........
 96: a  1.67774776
 97: a -1.05205570
 98: a -0.63690737
 99: a  0.56539163
100: a  0.38015779
```

## Joins

```{r}
DT1 <- data.table(x=c('a', 'a', 'b', 'dt1'), y=1:4); DT1
     x y
1:   a 1
2:   a 2
3:   b 3
4: dt1 4

DT2 <- data.table(x=c('a', 'b', 'dt2'), z=5:7); DT2
     x z
1:   a 5
2:   b 6
3: dt2 7
setkey(DT1, x); setkey(DT2, x)
merge(DT1, DT2)
   x y z
1: a 1 5
2: a 2 5
3: b 3 6

```

## Fast reading

```{r,cache=TRUE}
big_df <- data.frame(x=rnorm(1E6), y=rnorm(1E6))
file <- tempfile()
write.table(big_df, file=file, row.names=FALSE, col.names=TRUE, sep="\t", quote=FALSE)
system.time(fread(file))
# Read 1000000 rows and 2 (of 2) columns from 0.035 GB file in 00:00:04
   user  system elapsed 
   3.07    0.02    3.22 

system.time(read.table(file, header=TRUE, sep="\t"))
# user  system elapsed 
  20.31    0.11   20.58 

```

## Summary and further reading

* The latest development version contains new functions like `melt` and `dcast` for data.tables 
  * [https://r-forge.r-project.org/scm/viewvc.php/pkg/NEWS?view=markup&root=datatable](https://r-forge.r-project.org/scm/viewvc.php/pkg/NEWS?view=markup&root=datatable)
* Here is a list of differences between data.table and data.frame
  * [http://stackoverflow.com/questions/13618488/what-you-can-do-with-data-frame-that-you-cant-in-data-table](http://stackoverflow.com/questions/13618488/what-you-can-do-with-data-frame-that-you-cant-in-data-table)
* Notes based on Raphael Gottardo's notes [https://github.com/raphg/Biostat-578/blob/master/Advanced_data_manipulation.Rpres](https://github.com/raphg/Biostat-578/blob/master/Advanced_data_manipulation.Rpres), who got them from Kevin Ushey.