---
title: "Endmemo Function A-Fa"
author: "Juanjo Moreiras"
date: "Monday, July 28, 2014"
output: html_document

---
###------------------"apply"function-------------

## apply; lapply sapply
## apply() function applies a function to margins of an array or matrix.
## apply(x,margin,func, ...)
. x: array
. margin: subscripts, for matrix, 1 for row, 2 for column
. func: the function

```{r}
BOD    #R built-in dataset, Biochemical Oxygen Demand
  Time demand
1    1    8.3
2    2   10.3
3    3   19.0
4    4   16.0
5    5   15.6
6    7   19.8

## Sum up for each row:
apply(BOD,1,sum)
[1]  9.3 12.3 22.0 20.0 20.6 26.8

## Sum up for each column:
apply(BOD,2,sum)
  Time demand 
    22     89 

## Multipy all values by 10:
apply(BOD,1:2,function(x) 10 * x)
     Time demand
[1,]   10     83
[2,]   20    103
[3,]   30    190
[4,]   40    160
[5,]   50    156
[6,]   70    198

## Used for array, margin set to 1:
x <- array(1:9)
apply(x,1,function(x) x * 10)
[1] 10 20 30 40 50 60 70 80 90

## Two dimension array, margin can be 1 or 2:
x <- array(1:9,c(3,3)); x
     [,1] [,2] [,3]
[1,]    1    4    7
[2,]    2    5    8
[3,]    3    6    9

## lapply() function can handle data frame with similar results, return is a list:
lapply(BOD,sum)
$Time
[1] 22
$demand
[1] 89

lapply(BOD,mean)
$Time
[1] 3.666667
$demand
[1] 14.83333

## sapply() has similar function, it defines "simplify=TRUE" by default, thus return a vector:
sapply(BOD,sum)
  Time demand 
    22     89 
sapply(BOD,sum,simplify=FALSE)
$Time
[1] 22
$demand
[1] 89
```

###-----------------R "attach"" function----------------

## attach(x)
## x: dataframe, matrix, list

```{r}
setwd("~/DataScientist/RUcla/RUclaData")
x <- read.csv("anova.csv",header=T,sep=",")

## We can't use the variable "Gender" in R Search Path:
gender
## we have to use
x$gender
## After attach the object "x", "Gender" can be used globally:
attach(x)
gender
[1] m m m m m f m m f m m f m m m m f m m m m m m f m m m f m m m m f m m m m
 [38] m m m m m m m m m f m f m m m m m f m m f m m f m m m m f m m m m m m m m..............................
detach() function reverses the process:
>detach(x)
```

### ---------- "cbind"" function---------------

## cbind() function combines vector, matrix or data frame by columns.
cbind(x1,x2,...)
x1,x2:vector, matrix, data frames

## data1.csv:
A,m,-0.54
A,f,-0.8
B,f,-1.03
C,m,-0.41

## data2.csv:
32,New York
21,Houston
34,Seattle
67,Houston

```{r}
##Read in the data from the file:
x <- read.csv("data1.csv",header=T,sep=",")
x2 <- read.csv("data2.csv",header=T,sep=","); x3
x3 <- cbind(x,x2)
  Subtype Gender Expression Age     City
1       A      m      -0.54  32 New York
2       A      f      -0.80  21  Houston
3       B      f      -1.03  34  Seattle
4       C      m      -0.41  67  Houston

The row number of the two datasets must be equal.
```

###----------------"cut" function--------------

## cut() function divides a numeric vector into different ranges. 
## cut(x, breaks, labels = NULL,
    include.lowest = FALSE, right = TRUE, dig.lab = 3,
    ordered_result = FALSE, ...)
. x: numeric vector
. breaks: break points, number or numeric vector.
. labels: level labels, character vector.

```{r}
x <- stats::rnorm(100); x
> x
  [1] -0.154103462  0.271704132 -0.234160855  0.764474679  0.438237645
  [6] -0.763854668  1.303402711  0.051660328  1.064258570  0.079144697
 [11] -0.704381407  2.239763673 -0.749203152  0.601148921 -0.174814689
   ...............
[96] -2.025528709  0.949853574 -0.917568904 -1.103676434  0.728284402

## Divide the data into ranges -5 ~ 5:
c <- cut(x,breaks=-5:5); c
  [1] (-1,0]  (0,1]   (-1,0]  (0,1]   (0,1]   (-1,0]  (1,2]   (0,1]   (1,2]  
 [10] (0,1]   (-1,0]  (2,3]   (-1,0]  (0,1]   (-1,0]  (0,1]   (0,1]   (-1,0] 
 [19] (-2,-1] (0,1]   (0,1]   (-1,0]  (0,1]   (0,1]   (-1,0]  (-1,0]  (-1,0] .............
 [91] (1,2]   (-1,0]  (-1,0]  (-1,0]  (1,2]   (-3,-2] (0,1]   (-1,0]  (-2,-1]
[100] (0,1] 
10 Levels: (-5,-4] (-4,-3] (-3,-2] (-2,-1] (-1,0] (0,1] (1,2] (2,3] ... (4,5]

## Check the data distribution in different ranges:
summary(c)        #or table(c)
(-5,-4] (-4,-3] (-3,-2] (-2,-1]  (-1,0]   (0,1]   (1,2]   (2,3]   (3,4]   (4,5] 
      0       0       2      14      35      38      10       1       0       0 

## Now, the numbers are divided into 10 levels, the default step is 1. Some levels are empty. Let's try just define the total level number:
x <- stats::rnorm(100)           #random numbers, different every time
c <- cut(x,breaks=10,dig.lab=2)
summary(c)
(-2,-1.6]   (-1.6,-1.1]  (-1.1,-0.69] (-0.69,-0.24]  (-0.24,0.21] 
            5             5            13            20            18 
  (0.21,0.65]    (0.65,1.1]     (1.1,1.5]       (1.5,2]       (2,2.4] 
           12            14             6             3             4 

## Next, label all the levels:
x <- stats::rnorm(100)                 #random numbers, different every time
c <- cut(x,breaks=10,dig.lab=2,labels=1:10)
summary(c)
 1  2  3  4  5  6  7  8  9 10 
 5  5 13 20 18 12 14  6  3  4

## Try again, divide into different ranges (break points):
x <- stats::rnorm(100) #random numbers, different every time
c <- cut(x,breaks=c(-2,0,1,2))
table(c)
-2,0]  (0,1]  (1,2] 
    52     32     11 
```

###------------------"density" function--------------

## density() function computes kernel density estimates.
density(x, bw = "nrd0", adjust = 1,
        kernel = c("gaussian", "epanechnikov", "rectangular",
                   "triangular", "biweight",
                   "cosine", "optcosine"),
        weights = NULL, window = kernel, width,
        give.Rkern = FALSE,
        n = 512, from, to, cut = 3, na.rm = FALSE, ...)
x: number vector
bw: smoothing bandwidth

```{r}
## Let generate 100 numbers randomly:
x <- stats::rnorm(100); x
 [1] -0.154103462  0.271704132 -0.234160855  0.764474679  0.438237645
  [6] -0.763854668  1.303402711  0.051660328  1.064258570  0.079144697
.................
[91]  1.570993774 -0.876630726 -0.343492831 -0.480549452  1.494723381
 [96] -2.025528709  0.949853574 -0.917568904 -1.103676434  0.728284402

d <- density(x); d
Call:
      density.default(x = x)
Data: x (100 obs.);     Bandwidth 'bw' = 0.3184
       x                 y            
 Min.   :-3.7891   Min.   :0.0001413  
 1st Qu.:-2.0431   1st Qu.:0.0117442  
 Median :-0.2971   Median :0.0627054  
 Mean   :-0.2971   Mean   :0.1430424  
 3rd Qu.: 1.4489   3rd Qu.:0.2957362  
 Max.   : 3.1949   Max.   :0.4192181 

## Plot the density:
plot(density(x),xlim=c(-4,4),col="blueviolet")
```

###---------------"dim" function-------------

## dim() function gets or sets the dimension of a matrix, array or data frame.
dim(x)
x: array, matrix or data frame.

```{r}
BOD #R Biochemical Oxygen Demand Dataset
  Time demand
1    1    8.3
2    2   10.3
3    3   19.0
4    4   16.0
5    5   15.6
6    7   19.8

class(BOD)
[1] "data.frame"

dim(BOD) #get dimension
[1] 6 2

##Set dimension of a matrix:
x <- rep(1:20); x

##Set dimension to 2 × 5:
dim(x) <- c(2,10); x

```

###-------------"exp" function--------------

## exp(x) function compute the exponential value of a number or number vector, ex.

```{r}
x <- 5
exp(x)
[1] 148.4132

y <- rep(1:20)
exp(y)
         [,1]     [,2]     [,3]     [,4]      [,5]
[1,] 2.718282 20.08554 148.4132 1096.633  8103.084
[2,] 7.389056 54.59815 403.4288 2980.958 22026.466

## ^ operator calculates a raised to power b:
2^3
[1] 8

4 ^ (1/2)
[1] 2

```

###------------------"factor" function--------------

## Factors variable is a vector of categorical data. factor() function creates a factor variable, and calculates the categorical distribution of a vector data.

factor(x = character(), levels, labels = levels,
       exclude = NA, ordered = is.ordered(x))

x: a vector of data

```{r}
v <- c(1,3,5,8,2,1,3,5,3,5)
is.factor(v)
[1] FALSE

## Calculates the categorical distribution:
factor(v)
 [1] 1 3 5 8 2 1 3 5 3 5
Levels: 1 2 3 5 8

x <- factor(v); x
 [1] 1 3 5 8 2 1 3 5 3 5
Levels: 1 2 3 5 8

is.factor(x)
[1] TRUE

## Select levels:
x <- factor(v, levels=c(2,1)); x
 [1] 1    <NA> <NA> <NA> 2    1    <NA> <NA> <NA> <NA>
Levels: 2 1

## Change the level value:
levels(x) <- c("two","one"); x
 [1] one    <NA> <NA> <NA> two    one    <NA> <NA> <NA> <NA>
Levels: two one
```

 
 