---
title: "Endmemo Functions Sp-Z"
author: "Juanjo Moreiras"
date: "Monday, July 28, 2014"
output: html_document
---

### ---------------------"split" function-----------------

## split() function divides the data in a vector. unsplit() funtion do the reverse.

split(x, f, drop = FALSE, ...)
split(x, f, drop = FALSE, ...) <- value
unsplit(value, f, drop = FALSE)

x: vector, data frame
f: indices
drop: discard non existing levels or not

```{r}
## Let first read in the data from the file:
x <- read.csv("anova.csv",header=T,sep=",")

## Split the "Expression" values into two groups based on "Gender" variable, "f" for female group, and "m" for male group:
g <- split(x$Expression, x$Gender); g
$f
  [1] -0.66 -1.15 -0.30 -0.40 -0.24 -0.92  0.48 -1.68 -0.80 -0.55 -0.11 -1.26
 [13] -0.11  0.13  0.81  0.45  0.74 -0.31 -0.18 -0.08  0.54 -0.35  0.38 -0.39
 [25] .....................
[121]  0.97 -0.72 -0.04  0.38 -1.87 -2.09 -1.54  0.09 -0.25  0.51  0.33 -1.29
[133] -0.51 -0.50 -0.52
$m
  [1] -0.54 -0.80 -1.03 -0.41 -1.31 -0.43  1.01  0.14  1.42 -0.16  0.15 -0.62
 [13] -0.42 -0.35 -0.42  0.32 -0.57 -0.07 -0.06  0.02 -0.39 -0.74 -0.09 -0.03...
.......
[133] -0.39 -0.40  1.18 -0.30 -1.91 -1.14  0.13 -0.34 -0.44  0.52  1.11 -0.89
[145] -0.17 -1.62

## Calculate the length, mean value of each group:
sapply(g,length)
  f   m 
135 146 

sapply(g,mean)
         f          m 
-0.3946667 -0.2227397

## You may use lapply, return is a list:
lapply(g,mean)
$f
[1] -0.3946667
$m
[1] -0.2227397

## unsplit() function combines the groups:
unsplit(g,x$Gender)
```

###--------------"string" Functions----------------

## "string"" functions include substr(x), nchar(x), toupper(x), tolower(x), strsplit(x,y),paste(...), and regular expression functions sub(...), grep(...) etc.

```{r}
s <- "EndMemo.com R Language Tutorial"
substr(s,0,7)
[1] "EndMemo"

## Get string length: 
nchar(s)
[1] 31

## To uppercase:
x <- toupper(s); x
[1] "ENDMEMO.COM R LANGUAGE TUTORIAL"

##To lowercase:
x <- tolower(s); x
[1] "endmemo.com r language tutorial"

## Split the string at letter "o":
x <- strsplit(s,"o")
[[1]]
[1] "EndMem"           ".c"               "m R Language Tut" "rial"

## Concatenate two strings:
x <- paste(x," -- String Functions",sep=""); x
[1] "endmemo.com r language tutorial -- String Functions"

## Substring replacement:
x <- sub("Tutorial","Examples",s); x
[1] "EndMemo.com R Language Examples"

## Use regular expression:
>x <- sub("n.+e","XXX",s); x
>x
[1] "EXXX Tutorial"

## Please see grep() function for more regular expression handling of string.
```

### R "tapply" function

## tapply() applies a function to each cell of a ragged array.

tapply(X, INDEX, FUN = NULL, ..., simplify = TRUE)

. X: vector
. INDEX: list of one of more factors
. FUN: the function
```{r}    
Orange            #R built-in dataset, Growth of Orange Trees
   Tree  age circumference
1     1  118            30
2     1  484            58
3     1  664            87
4     1 1004           115
5     1 1231           120
.........
34    5 1372           174
35    5 1582           177

## Calculate the mean circumference of different Tree groups:
tapply(Orange$circumference,Orange$Tree,mean)
        3         1         5         2         4 
 94.00000  99.57143 111.14286 135.28571 139.28571 

# Return a list:
tapply(Orange$circumference,Orange$Tree,mean,simplify=FALSE)
$`3`
[1] 94
$`1`
[1] 99.57143
$`5`
[1] 111.1429
$`2`
[1] 135.2857
$`4`

[1] 139.2857
. simplify: if true, return an array of scalar, other wise an array of list
```

