---
title: "Epi Aragon basic"
author: "Juanjo Moreiras"
date: "Thursday, July 31, 2014"
output: html_document
---
---
title: "Epi Aragon Very Basic"
author: "Juanjo Moreiras"
date: "Thursday, July 31, 2014"
output: html_document
-----

##------------------EPIDEMIOLOGY ARAGON VERY BASIC EDA--------------------------

#_____IMPORT, EDIT, SORT, INDEX(SUBSET), TRANSFORM, MERG, MISSING ______

##__ 1. IMPORTING DATA FROM A FILE (pag. 115)__
#_ 1.1 READING AS ASCII TEXT DATA FILE

# Using University Group Diabetes Program randomized clinical trial text
# Data file is comma-delimite with row names and a header (ugdp.txt)

```{r Aragon}
setwd("~/DataScientist/RUcla/RUclaData")

ud <- read.table("http://www.medepi.net/data/ugdp.txt",
                 header = TRUE, sep = ",")
head(ud)

# The same data, but without row names and without header
cnames <- c("Status", "Treatment", "Agegrp")
udat2 <- read.table("http://www.medepi.net/data/ugdp2.txt",
                    + header = FALSE, sep = ",", col.names = cnames)
head(udat2)
# The same data file as it might appear as a ?x formatted file, (fwf)
# Needs a data dictionary to know which columns contains which fields

cnames <- c("Status", "Treatment", "Agegrp")
udat3 <- read.fwf("http://www.medepi.net/data/ugdp3.txt",
                  width = c(8, 11, 3), col.names = cnames, 
                  strip.white = TRUE)
head(udat3)

# The same data as a fixed width formatted file but with numeric codes
# Needs a data dictionary

cnames <- c("Status", "Treatment", "Agegrp")
udat4 <- read.fwf("http://www.medepi.net/data/ugdp4.txt",
                  width = c(1, 1, 1), col.names = cnames)
head(udat4)
```

#_ 1.2 READING DATA FROM BINARY FORMAT: e.g., STATA

```{r Aragon}

library(foreign)
# For example: 
idat <- read.dta("c:/.../data/infert.dta")
```

#_ 1.3 IMPORTING DATA USING URL
# loading the Western Collaborative Group Study from a web server

```{r Aragon}
wdat <- read.table("http://www.medepi.net/data/wcgs.txt",
                   header = TRUE, sep = ",")
str(wdat)
```

##__ 2. EDITING DATA __
#_ 2.1 EDIT OR FIX FUNCTIONS

```{r}

# Using WNV surveillance raw data as a data frame
wd <- read.table("http://www.medepi.net/data/wnv/wnv2004raw.txt",
                 header = TRUE, sep = ",", as.is = TRUE)
str(wd) 

# First assess in each variable, what needs to be "cleaned up"
lapply(wd, table) #apply ’table’ function to fields

# The results showed: we have 779 obs and variables with missing values
# Missing values need to be converted to NAs
wd[wd=="." | wd=="Unknown"] <- NA

# Evalute a variable to verify if missing values were converted to NAs
table(wd$death)     #(without NAs)
table(wd$death, exclude=NULL)   #(with NAs)
```

#_ 2.2 TEXT PROCESSING

```{r Aragon}
# e.g.,using "substr" fraction: "mm/dd/yy" to "yyyy-mm-dd"
bd <- c("07/17/96","12/09/00","11/07/97")
mon <- substr(bd, start=1, stop=2); mon
day <- substr(bd, 4, 5); day
yr <- as.numeric(substr(bd, 7, 8)); yr
yr2 <- ifelse(yr<=19, yr+2000, yr+1900); yr2
bdfin <- paste(yr2, "-", mon, "-", day, sep=""); bdfin

# e.g.,with "substr": “00” to “2000”, and “96”,“97 to “1996”,“1997”
bd
substr(bd, 3, 3) <- "-"
substr(bd, 6, 6) <- "-"
bd
```

##__ 3. SORTING DATA __
# Using sort function

```{r}
sort (x)
sort(x, decreasing = TRUE)
rev(sort(x))

# Using "order" function
# Sort one vector with ordering elements form another vector
x <- sample(1:20, 5); x
sort(x) #sorts as expected
y <- sample(1:20, 5); y
order(y)       #4th element to 1st position, 5th to 2nd, etc.
x[order(y)]    #use order(y) to sort elements of x
```

#_3.1 SORTING DATA FRAMES
# Creating a small data set

```{r}
sex <- rep(c("Male", "Female"), c(4, 4))
ethnicity <- rep(c("White", "African American", "Latino",
                   "Asian"), 2)
age <- sample(1:100, 8)
dat <- data.frame(age, sex, ethnicity)
dat <- dat[sample(1:8, 8),] #randomly order rows
dat

# Sort based on the ordering of one variable
dat[order(dat$age),]

# Sort based on the ordering of two variables
dat[order(dat$sex, dat$age),] 
```

##__ 4. INDEXING (SUBSETTING) DATA __

```{r}
# Using the Oswego foodborne illness dataset
odat <- read.table("http://www.medepi.net/data/oswego.txt",
                   header = TRUE, as.is = TRUE, sep = "")
str(odat)
```

#_4.1 INDEXING
# Indexing rows with cases
# First create a new data with only cases
# Generate a logical vector TRUE for every value of odat$ill

```{r}
cases <- odat$ill=="Y"
cases

# we extracted rows with cases by indexing the data with a logical vector 
odat.ca <- odat[cases, ]
odat.ca[, 1:8]

# e.g., Let’s create a data set with female cases
# age less than the median age, and consumed vanilla ice cream
fem.cases.vic <- odat$ill=="Y" & odat$sex=="F" &
  odat$vanilla.ice.cream=="Y" & odat$age < median(odat$age)
odat.fcv <- odat[fem.cases.vic, ]
odat.fcv[ , c(1:6, 19)]
```

#_4.2 SUBSETTING
# Using "subset" function
# Using data with female cases, age <median, and ate vanilla ice cream
# the ?rst argument is the data frame object name
# the second argument (subset) evaluates to a logical vector
# Third argument (slect) specifies the fields to keep

```{r}
odat.fcv <- subset(odat, subset = {ill=="Y" & sex=="F" &
                                     vanilla.ice.cream=="Y" & age < median(odat$age)},
                   select = c(id:onset.date, vanilla.ice.cream))  #there is an error
odat.fcv
```

##__ 5. TRANSFORMING DATA __
#_5.1 CREATING CATEGORICAL VARIABLES (FACTORS)

```{r}
# Using the Oswego foodborne illness dataset
odat <- read.table("http://www.medepi.net/data/oswego.txt",
                   header = TRUE, as.is = TRUE, sep = "")

# create new variable with seven age categories
# < 1, 1 to 4, 5 to 14, 15 to 24, 25 to 44, 45 to 64, and 65+
# Using several methods

# Using "cut" function
agecat <- cut(odat$age, breaks = c(0, 1, 5, 15, 25, 45, 65, 100))
agecat

# The notation (15,25] means open left boundary (>15), closed right (<=25)
# It is better use closed left and open right boundaries [15,25)
agecat <- cut(odat$age, breaks = c(0, 1, 5, 15, 25, 45,
                                   65, 100), right = FALSE)
agecat

# Adding labels since we are notfamiliar with open and closed notations
agelabs <- c("<1", "1-4", "5-14", "15-24", "25-44", "45-64","65+")
agecat <- cut(odat$age, breaks = c(0, 1, 5, 15, 25, 45,
                                   65, 100), right = FALSE, labels = agelabs)
agecat
table(agecat, case = odat$ill)
```

#_ 5.2 RE-CODING LEVELS OF A CATEGORICAL VARIABLE

# keeping track of integer codes for categorical variables is unnecessary
# however, if we feel the need to do so, here’s how.

```{r}
# Create categorical variable
ethlabs <- c("White", "Black", "Latino", "Asian")
ethnicity <- sample(ethlabs, 100, replace = T)
ethnicity <- factor(ethnicity, levels = ethlabs)
ethnicity
#[1] White  White  Latino White  Latino Asian  White  Asian  White  Black 
# [11] Black  Asian  White  Asian  Latino Latino Latino White  Black  White 
# [21] White  Latino White  Black  Latino Asian  Asian  White  White  Latino
# [31] Black  Black  Black  Black  Latino Latino Asian  Asian  White  Latino
# [41] Black  Asian  Latino Latino Black  White  Black  White  Black  Black 
# [51] Asian  Latino Latino Black  Latino Latino Latino Latino Asian  Asian 
# [61] Latino Black  Latino White  Latino White  Black  White  White  White 
# [71] Asian  White  Black  Black  White  Black  Latino Latino Latino Latino
# [81] Black  Black  Asian  Black  White  Black  Latino White  Black  Latino
# [91] Asian  Black  Asian  Asian  Asian  Latino Asian  Latino White  Black 
# Levels: White Black Latino Asian

# The first level option is the reference level 
#Displaying the numeric code with "unclass" function
x <- unclass(ethnicity); x

# To recover the original factor,
factor(x,labels=levels(x))
# [1] White  White  Latino White  Latino Asian  White  Asian  White  Black 
# [11] Black  Asian  White  Asian  Latino Latino Latino White  Black  White 
# [21] White  Latino White  Black  Latino Asian  Asian  White  White  Latino
# [31] Black  Black  Black  Black  Latino Latino Asian  Asian  White  Latino
# [41] Black  Asian  Latino Latino Black  White  Black  White  Black  Black 
# [51] Asian  Latino Latino Black  Latino Latino Latino Latino Asian  Asian 
# [61] Latino Black  Latino White  Latino White  Black  White  White  White 
# [71] Asian  White  Black  Black  White  Black  Latino Latino Latino Latino
# [81] Black  Black  Asian  Black  White  Black  Latino White  Black  Latino
# [91] Asian  Black  Asian  Asian 

# we can re-label the levels using the "levels" function
# and assigning to it a character vector of new labels
# Make sure the order of the new labels corresponds toorder of the factor levels 
levels(ethnicity2) <- c("Caucasion", "African American",
                        "Hispanic", "Asian")
table(ethnicity2)

# re-order and re-label at the same time using the levels function
table(ethnicity)
# ethnicity
# White  Black Latino  Asian 
#25     26     30     19 
ethnicity3 <- ethnicity
levels(ethnicity3) <- list(Hispanic = "Latino", Asian = "Asian",
                           Caucasion = "White", "African American" = "Black")
table(ethnicity3)
# ethnicity3
# Hispanic            Asian        Caucasion African American 
#  30               19               25               26 

# Sorting the factor levels by using the factor function
table(ethnicity)
ethnicity
# White Black Latino Asian
# 23 31 28 18
ethnicity5a <- factor(ethnicity, sort(levels(ethnicity)))
table(ethnicity5a)
ethnicity5a
# ethnicity5a
# Asian  Black Latino  White 
#19     26     30     25
```

5.3_ SETTING FACTOR REFERENCE LEVEL

```{r}
# Using the "relevel" function
levels(ethnicity)
# [1] "White" "Black" "Latino" "Asian"
ethnicity6 <- relevel(ethnicity, ref = "Asian")
levels(ethnicity6)
# [1] "Asian" "White" "Black" "Latino"
```

5.4_ FACTORS INSTEAD OF DUMMY VARIABLES
# it is unnecessary to create dummy variables
# create a factor with number of levels and set the reference level.
-------
  
5.5_ CONDITIONALLY TRANSFORMING ELEMENTS OF A VECTOR

```{r}
# Using "ifelse" factor
# works as: ifelse(test, if test = TRUE do this, else do this)
x <- sample(c("M", "F"), 10, replace = T); x
# [1] "M" "F" "M" "F" "M" "F" "M" "F" "M" "F"

y <- ifelse(x=="M", "Male", "Female"); y
# [1] "Male" "Female" "Male" "Female" "Male" "Female"
# [7] "Male" "Female" "Male" "Female"
```

##6. __ MERGING DATA __

# Usually not necessary to merge data objects into one data object
# On occasion is necessary to merge two data frames into one data frames

# Look two related data frames that are linked by an indexing ?eld
# First data frame contains telephones for 5 employees and fname is the indexing ?eld

# For example:
# phone
#   fname phonenum phonetype
# 1 Tomas 643-4935 work
# 2 Tomas 847-9139 mobile
# 3 Tomas 643-4926 fax
# 4 Chris 643-3932 work
# 5 Chris 643-4926 fax
# 6 Wayne 643-4934 work
# 7 Wayne 643-4926 fax
# 8 Ray 643-4933 work
# 9 Ray 643-4926 fax
# 10 Diana 643-3931 work

#The second data are email for 3 employees and name is the indexing ?eld
# > email
#   name mail mailtype
# 1 Tomas aragon@berkeley.edu Work
# 2 Tomas aragon@medepi.net Personal
# 3 Wayne enanoria@berkeley.edu Work
# 4 Wayne enanoria@idready.org Work
# 5 Chris csiador@berkeley.edu Work
# 6 Chris csiador@yahoo.com Personal

```{r}
# To merge these two data frames use the "merge" function
dat <- merge(email, phone, by.x="name", by.y="fname")
dat
# name mail mailtype phonenum phonetype
# 1 Chris csiador@berkeley.edu Work 643-3932 work
# 2 Chris csiador@yahoo.com Personal 643-3932 work
# 3 Chris csiador@berkeley.edu Work 643-4926 fax
# 4 Chris csiador@yahoo.com Personal 643-4926 fax
# 5 Tomas aragon@berkeley.edu Work 643-4935 work
# 6 Tomas aragon@medepi.net Personal 643-4935 work
# 7 Tomas aragon@berkeley.edu Work 847-9139 mobile
# 8 Tomas aragon@medepi.net Personal 847-9139 mobile
# 9 Tomas aragon@berkeley.edu Work 643-4926 fax
# 10 Tomas aragon@medepi.net Personal 643-4926 fax
# 11 Wayne enanoria@berkeley.edu Work 643-4934 work
# 12 Wayne enanoria@idready.org Work 643-4934 work
# 13 Wayne enanoria@berkeley.edu Work 643-4926 fax
# 14 Wayne enanoria@idready.org Work 643-4926 fax

dat <- merge(phone, email, by.x="fname", by.y="name")
dat
# fname phonenum phonetype mail mailtype
# 1 Chris 643-3932 work csiador@berkeley.edu Work
# 2 Chris 643-4926 fax csiador@berkeley.edu Work
# 3 Chris 643-3932 work cvsiador@yahoo.com Personal
# 4 Chris 643-4926 fax cvsiador@yahoo.com Personal
# 5 Tomas 643-4935 work aragon@berkeley.edu Work
# 6 Tomas 847-9139 mobile aragon@berkeley.edu Work
# 7 Tomas 643-4926 fax aragon@berkeley.edu Work
# 8 Tomas 643-4935 work aragon@medepi.net Personal
# 9 Tomas 847-9139 mobile aragon@medepi.net Personal
# 10 Tomas 643-4926 fax aragon@medepi.net Personal
# 11 Wayne 643-4934 work enanoria@berkeley.edu Work
# 12 Wayne 643-4926 fax enanoria@berkeley.edu Work
# 13 Wayne 643-4934 work enanoria@idready.org Work
# 14 Wayne 643-4926 fax enanoria@idready.org Work

# The by.x and by.y options identify the indexing ?elds
# To merge the union of the indexing ?elds, set all=TRUE
dat <- merge(phone, email, by.x="fname", by.y="name",
             all=TRUE)
dat
# 1 Chris 643-3932 work csiador@berkeley.edu Work
# 2 Chris 643-4926 fax csiador@berkeley.edu Work
# 3 Chris 643-3932 work csiador@yahoo.com Personal
# 4 Chris 643-4926 fax csiador@yahoo.com Personal
# 5 Diana 643-3931 work <NA> <NA>
# 6 Ray 643-4933 work <NA> <NA>
# 7 Ray 643-4926 fax <NA> <NA>
# 8 Tomas 643-4935 work aragon@berkeley.edu Work
# 9 Tomas 847-9139 mobile aragon@berkeley.edu Work
# 10 Tomas 643-4926 fax aragon@berkeley.edu Work
# 11 Tomas 643-4935 work aragon@medepi.net Personal
# 12 Tomas 847-9139 mobile aragon@medepi.net Personal
# 13 Tomas 643-4926 fax aragon@medepi.net Personal
# 14 Wayne 643-4934 work enanoria@berkeley.edu Work
# 15 Wayne 643-4926 fax enanoria@berkeley.edu Work
# 16 Wayne 643-4934 work enanoria@idready.org Work
# 17 Wayne 643-4926 fax enanoria@idready.org Work
```

##__ 7. MISSING AND "NOT AVAILABLE" VALUES (page 142)

```{r}

# NAs not always represents a missing value
# Objects tha contain NA return NA
# Removing NAs with na.rm option
sum(x) # answer NA
Mean(x) # answer NA
sum(x, na.rm = TRUE)  # better
mean(x, na.rm = TRUE) # better
```

#_ 7.1 TESTING, INDEXING, REPLACING, RECODING

```{r}
# using is.na generateS a logical vector to identify positions OD NAs
x <- c(10, NA, 33, NA, 57)
y <- c(NA, 24, NA, 47, NA)

is.na(x)               #generate logical vector
# [1] FALSE TRUE FALSE TRUE FALSE

which(is.na(x))        #which positions are NA
# [1] 2 4

x[!is.na(x)]           #index original vector
[1] 10 33 57

y[is.na(x)]            #index other vector
[1] 24 47

x[is.na(x)] <- 999     #replacement
x
#[1] 10 999 33 999 57

# recoding missing values to NA with replacement
x <- c(1, -99, 3, -88, 5)
x[x==-99 | x==-88] <- NA
x
#[1] 1 NA 3 NA 5

#Recoding missing values to NA using replacement
fname <- c("Tom", "Unknown", "Jerry")
age <- c(56, 34, -999)
z1 <- z2 <- data.frame(fname, age)
z1
#  fname age
# 1 Tom 56
# 2 Unknown 34
# 3 Jerry -999

# Replacement one column at a time
z1$fname[z1$fname=="Unknown"] <- NA
z1$age[z1$age==-999] <- NA
z1
#  fname age
# 1 Tom 56
# 2 <NA> 34
# 3 Jerry NA

# Global replacement
z2[z2=="Unknown" | z2==-999] <- NA
z2
#  fname age
# 1 Tom 56
# 2 <NA> 34
# 3 Jerry NA
```

#_ 7.2 IMPORTING MISSING VALUES WITH READ.TABLE FUNCTION

# The default setting is na.strings="NA"
# na.strings option to specify what characters are to be converted to NA
# Blank ?elds are also considered to be missing values

```{r}
# Suppose the data contains 999, 888, and . to represent missing values

mydat <- read.table("dataset.txt", na.strings = c(999, 888, "."))

# If a number, say 999, represents a missing value in one ?eld but a valid value in
another ?eld
# import the data using the as.is=TRUE option
# Then replace missing values in the data one ?eld at a time, 
# And convert categorical variables to factors
```

## 8__ DATES AND TIMES __  (some examples)
# 8.1_ AS.DATE FUNCTION

```{r}

# as.Date function converts calendar dates ino a Date objects
bdays <- c("11/2/1959", "1/1/1970")
bdays
# "11/2/1959" "1/1/1970"

#convert to Julian dates
bdays.julian <- as.Date(bdays, format = "%m/%d/%Y")
bdays.julian
#"1959-11-02" "1970-01-01"

#Although this looks a character vectors. it is class "Date"
#display Julian dates
as.numeric(bdays.julian)
# [1] -3713 0

#calculate age as of today’s date
date.today <- Sys.Date()
date.today
#[1] "2005-09-25"

age <- (date.today - bdays.julian)/365.25
age

#Time differences of 45.89733, 35.73169 days
#the display of ’days’ is not correct
#truncate number to get "age"
age2 <- trunc(as.numeric(age))
age2
# [1] 45 35

## create date frame
bd <- data.frame(Birthday = bdays, Standard = bdays.julian,
                 Julian = as.numeric(bdays.julian), Age = age2)
bd
#   Birthday Standard Julian Age
#1 11/2/1959 1959-11-02 -3713 45
#2 1/1/1970 1970-01-01 0 35 

# Selected format options; for a complete list see help(strptime).
"%a" Abbreviated weekday name.
"%A" Full weekday name.
"%b" Abbreviated month name.
"%B" Full month name.
"%d" Day of the month as decimal number (01-31)
"%j" Day of year as decimal number (001-366).
"%m" Month as decimal number (01-12).
"%U" Week of the year as decimal number (00-53) using the
first Sunday as day 1 of week 1.
"%w" Weekday as decimal number (0-6, Sunday is 0).
"%W" Week of the year as decimal number (00-53) using the
first Monday as day 1 of week 1.
"%y" Year without century (00-99). If you use this on input,
which century you get is system-specific. So don’t!
  Often values up to 69 (or 68) are prefixed by 20 and
70-99 by 19.
"%Y" Year with century.

# Some examples of converting dates with different formats:
as.Date("November 2, 1959", format = "%B %d, %Y")
# [1] "1959-11-02"
# as.Date("11/2/1959", format = "%m/%d/%Y")
# [1] "1959-11-02"
#caution using 2-digit year
as.Date("11/2/59", format = "%m/%d/%y")
# [1] "2059-11-02"
as.Date("02Nov1959", format = "%d%b%Y")
# [1] "1959-11-02"
# caution using 2-digit year
as.Date("02Nov59", format = "%d%b%y")
# [1] "2059-11-02"
# standard format does not require format option
as.Date("1959-11-02")
# [1] "1959-11-02"

#The weekdays, months, quarters, julian functions
mydates <- c("2004-01-15","2004-04-15","2004-10-15")
mydates <- as.Date(mydates)
weekdays(mydates)
# [1] "Thursday" "Thursday" "Friday"
months(mydates)
# [1] "January" "April" "October"
quarters(mydates)
# [1] "Q1" "Q2" "Q4"
julian(mydates)
# [1] 12432 12523 12706
```







































