---
title: "Epi Aragon""
author: "Juanjo Moreiras"
date: "Thursday, July 31, 2014"
output: html_document

```
##-----EPIDEMIOLOGY BASICS --------##

```{r}
install.packages("Epi")
library(Epi)
data( births )
str(births)
summary(births)
names(births)
summary(births$hyp)
```

## PARTS OF THE DATA FRAME

```{r aragon}
births[1,"bweight"]
births[2,"bweight"]
births[1:10, "bweight"]
births[, "bweight"]
```

## TURNING A VARIABLE INTO A FACTOR

```{r Aragon}
hyp <- factor(births$hyp)
str(births)
objects()
births <- transform(births, hyp=factor(hyp))
str(births)
births <- transform( births, hyp=factor(hyp,labels=c("normal","hyper")) )
str(births)
```

## FREQUENCY TABLES

```{r Aragon}
stat.table(hyp,data=births)
stat.table(sex,data=births)
stat.table(list(hyp,sex),data=births)
```

## GROUPING VALUES OF A METRIC VARIABLE
# Cutting matage into groups 20{29, 30{34, 35{39, 40{44 creating a factor called agegrp

```{r Aragon}
births <- transform(births,agegrp=cut(matage, breaks=c(20,30,35,40,45),right=FALSE))
stat.table(agegrp,data=births)
```

# Now, cutting a variable into a given number of intervals of equal length 

```{r Aragon}
births <- transform(births,agegrp=cut(matage,breaks=5,right=FALSE))
stat.table(agegrp,data=births)
```

## TABLES OF MEANS AND OTHER FUNCTIONS
# mean bweight by sex

```{r Aragon}
stat.table(sex, mean(bweight), data=births)
stat.table(sex,list("Mean birth weight"=mean(bweight)),data=births)
```

## 2x2 table of mean birth weight by sex and hypertension

```{r Aragon}
stat.table(list(sex,hyp),mean(bweight),data=births)
```

# Available functions: count, mean, weighted mean, sum, min, max, quantile,median, IQR, ratio, rates, odds

```{r Aragon}
stat.table(list(sex,hyp),list(count(),mean(bweight)),data=births)
stat.table(hyp, list("odds"=ratio(lowbw,1-lowbw,100)),data=births)
```

# Margins can be added to the tables, as required

```{r Aragon}
stat.table(sex, mean(bweight),data=births,margins=TRUE)
stat.table(list(sex,hyp),mean(bweight),data=births,margins=c(TRUE,FALSE))
```

## GENERATING NEW VARIABLES

```{r Aragon}
births <- transform(births,
                    num1=1,
                    num2=2,
                    logbw=log(bweight))
```

# Logical variables

```{r Aragon}
births <- transform(births, low=bweight<2000)
str(births)
```

# Restricting a command to a subset of the data.Example: list values bweight for hypertensive women

```{r Aragon}
births$bweight[births$hyp=="hyper"]
```

# Subsetting a data frame

```{r Aragon}
subset(births, hyp=="hyper")
```

## ATTACHING A DATA SET
# for typing "just" hyp instead the cumbersome command "births$hyp"
# Places a copy of the variable in the births data in position 2

```{r Aragon}
attach(births)
objects(2)
summary(hyp)

# do not forget to type detach
detach (births) 
search()  #be sure "package:Epi" is in place number two
```

## SOME TYPES OF PLOTS

```{r Aragon}
## Basic
hist(bweight, col="gray", border="white")
plot(gestwks, bweight, pch=16, col="green")
points(gestwks, bweight)

## More complicated
plot(gestwks, bweight, type="n")
points(gestwks[sex==1], bweight[sex==1], col="blue")
points(gestwks[sex==2], bweight[sex==2], col="red")

## Legends
legend("topleft", pch=1, legend=c("Boys","Girls"), col=c("blue","red"))
title("Birth weight vs gestational weeks in 500 singleton births")

## Using indexing 
c("blue","red")
 sex
c("blue","red")[sex]
plot( gestwks, bweight, pch=16, col=c("blue","red")[sex] )

## Separates mothers >40
oldmum <- ( matage >= 40 ) + 1
plot( gestwks, bweight, pch=c(16,3)[oldmum], col=c("blue","red")[sex] )
plot( gestwks, bweight, pch=c(16,3)[(matage>=40 )+1], col=c("blue","red")[sex] )

## Interacting with plots
plot( gestwks, bweight, pch=c(16,3)[(matage>=40 )+1], col=c("blue","red")[sex] )
identify( gestwks, bweight )

# Saving visualization
# file -> save as-> .pdf or into for example .emf

# Controling the plot
win.metafile(file="plot1.emf", height=3, width=4)
plot(gestwks, bweight)
dev.off()

# Manipulating the plot
# help->Html help->Packages->base->P->par
par( mfrow=c(2,3) )
par( mfrow=c(1,1) )
```

## THE FUNCTION EFFX

```{r Aragon}
library(Epi)
data(births)
help(births)

# First transforming some variables
births$hyp <- factor(births$hyp,labels=c("normal","hyper"))
births$sex <- factor(births$sex,labels=c("M","F"))
births$agegrp <- cut(births$matage,breaks=c(20,25,30,35,40,45),right=FALSE)
births$gest4 <- cut(births$gestwks,breaks=c(20,35,37,39,45),right=FALSE)

# The effect of sex on birth weight, measured as a difference in means,
effx(response=bweight,typ="metric",exposure=sex,data=births)
stat.table(sex,mean(bweight), data=births)

# The effect of sex on lowbw, measured as an odds ratio, is 1.43
effx(response=lowbw,typ="binary",exposure=sex,data=births)

# Factors on more than two levels
# There are 3 effects
effx(response=bweight,typ="metric",exposure=gest4,data=births)
stat.table(gest4,mean(bweight),data=births)

# Stratified effects
# The eects of hyp in the dierent strata dened by sex are
effx(bweight, type="metric", exposure=hyp, strata=sex,data=births)

# Controlling the effect of a variable for a third variable (potential confounder)
effx(bweight, type="metric", exposure=hyp, control=sex,data=births)

# Numeric esposures metric
effx(response=bweight, type="metric", exposure=gestwks,data=births)
effx(response=lowbw, type="binary", exposure=gestwks,data=births)

# Checking on linearity with a plot
with(births, plot(gestwks,bweight))
```

## DATES IN R

```{r}

data(diet)
str(diet) 

## Converting a character string (or a character variable) to date format
as.Date( "14/07/1952", format="%d/%m/%Y" )
as.numeric( as.Date( "14/07/1952", format="%d/%m/%Y" ) )

## Entering a fixed date. For example to terminate a follow-up 1st April 1975
newx <- pmin( diet$dox, as.Date( "1975-4-1", format="%F" ) )
   
## You can print dates in the format you like by using the function format.Date()
bdat <- as.Date( "1952-7-14", format="%F" )
format.Date( bdat, format="%A %d %B %Y" )

## The function cal.yr converts dates to fractional years
as.Date( "1952-7-14" )
 cal.yr( as.Date("1952-7-14") )
 cal.yr( "1952-7-14" )

## cal.yr can find all date-variables in the dataframe and convert them
data( diet )
str( diet )
str( cal.yr(diet) )