---
title: "Functions"
author: "Juanjo Moreiras"
date: "Friday, July 25, 2014"
output: pdf_document
---
## Summary
The things to consider when choosing an apply function are basically:

What class is my input data? - vector, matrix, data frame.
On which subsets of that data do I want the function to act? - rows, columns, all values.
What class will the function return? How is the original data structure transformed?

* apply             Apply Functions Over Array Margins
* by                Apply a Function to a Data Frame Split by Factors
* eapply            Apply a Function Over Values in an Environment
* lapply            Apply a Function over a List or Vector
* mapply            Apply a Function to Multiple List or Vector Arguments
* rapply            Recursively Apply a Function to a List
* tapply            Apply a Function Over a Ragged Array

-------------------

## 1. apply

# Description: "Returns a vector or array or list of values obtained by applying a function to margins of an array or matrix."

# apply(X, MARGIN, FUN, ...)

# X is an array or matrix
# MARGIN is a variable that determines whether the function is applied over rows (MARGIN=1), columns (MARGIN=2), or both (MARGIN=c(1,2));
# FUN is the function to be applied.

In essence, the apply function allows us to make entry-by-entry changes to data frames and matrices. If MARGIN=1, the function accepts each row of X as a vector argument, and returns a vector of the results. Similarly, if MARGIN=2 the function acts on the  columns of X. Most impressively,  when MARGIN=c(1,2) the function is applied to every entry of X. As for the FUN argument, this can be anything from a standard R function, such as sum or mean, to a custom function like translate above.

# An illustrative example

```{r}
## create a matrix of 10 rows x 2 columns
m <- matrix(c(1:10, 11:20), nrow = 10, ncol = 2); m
#     [,1] [,2]
# [1,]    1   11
# [2,]    2   12
# [3,]    3   13
# [4,]    4   14
# [5,]    5   15
# [6,]    6   16
# [7,]    7   17
# [8,]    8   18
# [9,]    9   19
# [10,]   10   20

# mean of the rows
apply(m, 1, mean)
# [1]  6  7  8  9 10 11 12 13 14 15

# mean of the columns
apply(m, 2, mean)
# [1]  5.5 15.5

# divide all values by 2
apply(m, 1:2, function(x) x/2)
#      [,1] [,2]
# [1,]  0.5  5.5
# [2,]  1.0  6.0
# [3,]  1.5  6.5
# [4,]  2.0  7.0
# [5,]  2.5  7.5
# [6,]  3.0  8.0
# [7,]  3.5  8.5
# [8,]  4.0  9.0
# [9,]  4.5  9.5
#[10,]  5.0 10.0
```

## 2. by

# Description: "by"" provides a way to split your data by factors and do calculations on each subset. It returns an object of class "by" and there are many, more complex ways to use it.

The by function is a little more complex. The documentation tells us that "a data frame is split by row into data frames subsetted by the values of one or more factors, and function 'FUN' is applied to each subset in turn." So, we use this one where factors are involved.

# An illustrative example with R dataset "iris" that contains of flower measurements

```{r}
attach(iris)
head(iris)
#Sepal.Length Sepal.Width Petal.Length Petal.Width Species
#1          5.1         3.5          1.4         0.2  setosa
#2          4.9         3.0          1.4         0.2  setosa
#3          4.7         3.2          1.3         0.2  setosa
#4          4.6         3.1          1.5         0.2  setosa
#5          5.0         3.6          1.4         0.2  setosa
#6          5.4         3.9          1.7         0.4  setosa

# get the mean of the first 4 variables, by species
by(iris[, 1:4], Species, colMeans)
#Species: setosa
#Sepal.Length  Sepal.Width Petal.Length  Petal.Width 
#       5.006        3.428        1.462        0.246 
#------------------------------------------------------------------ 
#Species: versicolor
#Sepal.Length  Sepal.Width Petal.Length  Petal.Width 
#       5.936        2.770        4.260        1.326 
#------------------------------------------------------------------ 
#Species: virginica
#Sepal.Length  Sepal.Width Petal.Length  Petal.Width 
#       6.588        2.974        5.552        2.026 
```

## 3. eapply

# Description: "eapply applies FUN to the named values from an environment and returns the results as a list."

This one is a little trickier, since you need to know something about environments in R. An environment, as the name suggests, is a self-contained object with its own variables and functions. To continue using our very simple example.

We don't often create our own environments, but they're commonly used by R packages, so it's good to know how to handle them. 

# An illustrative example

```{r}
# a new environment
e <- new.env()

# two environment variables, a and b
e$a <- 1:10
e$b <- 11:20

# mean of the variables
eapply(e, mean)
#$a
#[1] 5.5
#$b
#[1] 15.5
```

## 3. lapply

# Description: "lapply returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X."

which makes lapply one of the easier apply functions to understand.

# An illustrative example


```{r}
# create a list with 2 elements
l <- list(a = 1:10, b = 11:20); l
#$a
# [1]  1  2  3  4  5  6  7  8  9 10
#$b
# [1] 11 12 13 14 15 16 17 18 19 20

## the mean of the values in each element
lapply(l, mean)
#$a
#[1] 5.5
#$b
#[1] 15.5

# the sum of the values in each element
lapply(l, sum)
#$a
#[1] 55
#$b
[#1] 155
  
```

## 4.1 sapply

# Description: "sapply is a user-friendly version of lapply by default returning a vector or matrix if appropriate."
  
 That simply means that if lapply would have returned a list with elements $a and $b, sapply will return either a vector, with elements [['a']] and [['b']], or a matrix with column names "a" and "b". 
 
# Returning to our previous example

```{r}
## create a list with 2 elements
l <- list(a = 1:10, b = 11:20)
#$a
# [1]  1  2  3  4  5  6  7  8  9 10
#$b
# [1] 11 12 13 14 15 16 17 18 19 20

# mean of values using sapply
l.mean <- sapply(l, mean); l.mean
#   a    b 
# 5.5 15.5 

# what type of object was returned?
class(l.mean)
# [1] "numeric"

# it's a numeric vector, so we can get element "a" like this
l.mean[['a']]
#[1] 5.5
```

## 4.2 vapply

# Description: "vapply is similar to sapply, but has a pre-specified type of return value, so it can be safer (and sometimes faster) to use."

A third argument is supplied to vapply, which you can think of as a kind of template for the output. The documentation uses the fivenum function as an example, so let's go with that:

# vapply returned a matrix, where the column names correspond to the original list elements and the row names to the output template. Nice.

# An illustrative example

```{r}
l <- list(a = 1:10, b = 11:20); l
#$a
# [1]  1  2  3  4  5  6  7  8  9 10
#$$b
# [1] 11 12 13 14 15 16 17 18 19 20

## fivenum of values using vapply
l.fivenum <- vapply(l, fivenum, c(Min.=0, "1st Qu."=0, Median=0, "3rd Qu."=0, Max.=0))
class(l.fivenum)
#[1] "matrix"

## let's see it
l.fivenum
           a    b
Min.     1.0 11.0
1st Qu.  3.0 13.0
Median   5.5 15.5
3rd Qu.  8.0 18.0
Max.    10.0 20.0

```

## 4.3 replicate

# Description: "replicate is a wrapper for the common use of sapply for repeated evaluation of an expression (which will usually involve random number generation)."

The replicate function is very useful. Give it two mandatory arguments: the number of replications and the function to replicate; a third optional argument, simplify = T, tries to simplify the result to a vector or matrix. 

# An example - let's simulate 10 normal distributions, each with 10 observations:

```{r}
replicate(10, rnorm(10))
#             [,1]        [,2]        [,3]       [,4]        [,5]         [,6]
# [1,]  0.67947001 -1.94649409  0.28144696  0.5872913  2.22715085 -0.275918282
# [2,]  1.17298643 -0.01529898 -1.47314092 -1.3274354 -0.04105249  0.528666264
# [3,]  0.77272662 -2.36122644  0.06397576  1.5870779 -0.33926083  1.121164338
# [4,] -0.42702542 -0.90613885  0.83645668 -0.5462608 -0.87458396 -0.723858258
# [5,] -0.73892937 -0.57486661 -0.04418200 -0.1120936  0.08253614  1.319095242
# [6,]  2.93827883 -0.33363446  0.55405024 -0.4942736  0.66407615 -0.153623614
# [7,]  1.30037496 -0.26207115  0.49818215  1.0774543 -0.28206908  0.825488436
# [8,] -0.04153545 -0.23621632 -1.01192741  0.4364413 -2.28991601 -0.002867193
# [9,]  0.01262547  0.40247248  0.65816829  0.9541927 -1.63770154  0.328180660
#[10,]  0.96525278 -0.37850821 -0.85869035 -0.6055622  1.13756753 -0.371977151
#             [,7]        [,8]       [,9]       [,10]
# [1,]  0.03928297  0.34990909 -0.3159794  1.08871657
# [2,] -0.79258805 -0.30329668 -1.0902070  0.73356542
# [3,]  0.10673459 -0.02849216  0.8094840  0.06446245
# [4,] -0.84584079 -0.57308461 -1.3570979 -0.89801330
# [5,] -1.50226560 -2.35751419  1.2104163  0.74650696
# [6,] -0.32790991  0.80144695 -0.0071844  0.05742356
# [7,]  1.36719970  2.34148354  0.9148911  0.20451421
# [8,] -0.51112579 -0.53658159  1.5194130 -0.94250069
# [9,]  0.52017814 -1.22252527  0.4519702  0.08779704
#[10,]  1.35908918  1.09024342  0.5912627 -0.20709053

```

## 5. mapply

# Description: "mapply is a multivariate version of sapply. mapply applies FUN to the first elements of each (.) argument, the second elements, the third elements, and so on."

# The mapply documentation is full of quite complex examples, but here's a simple one.

Here, we sum l1$a[1] + l1$b[1] + l2$c[1] + l2$d[1] (1 + 11 + 21 + 31) to get 64, the first element of the returned list. All the way through to l1$a[10] + l1$b[10] + l2$c[10] + l2$d[10] (10 + 20 + 30 + 40) = 100, the last element.

```{r}
l1 <- list(a = c(1:10), b = c(11:20)); l1
#$a
# [1]  1  2  3  4  5  6  7  8  9 10
#$b
# [1] 11 12 13 14 15 16 17 18 19 20

l2 <- list(c = c(21:30), d = c(31:40)); l2
#$c
# [1] 21 22 23 24 25 26 27 28 29 30
#$d
# [1] 31 32 33 34 35 36 37 38 39 40

# sum the corresponding elements of l1 and l2
mapply(sum, l1$a, l1$b, l2$c, l2$d)
# [1]  64  68  72  76  80  84  88  92  96 100
```

## 6. rapply

# Description: "rapply is a recursive version of lapply."
 
We think "recursive" is a little misleading. What rapply does is apply functions to lists in different ways, depending on the arguments supplied.

# Consider that, the output of rapply depends on both the function and the how argument. When how = "list" (or "replace"), the original list structure is preserved. Otherwise, the default is to unlist, which results in a vector.

#You can also pass a "classes=" argument to rapply. For example, in a mixed list of numeric and character variables, you could specify that the function act only on the numeric values with "classes = numeric".

# Best illustrated by examples

```{r}
# let's start with our usual simple list example
l <- list(a = 1:10, b = 11:20)

# log2 of each value in the list
rapply(l, log2)
      a1       a2       a3       a4       a5       a6       a7       a8 
0.000000 1.000000 1.584963 2.000000 2.321928 2.584963 2.807355 3.000000 
      a9      a10       b1       b2       b3       b4       b5       b6 
3.169925 3.321928 3.459432 3.584963 3.700440 3.807355 3.906891 4.000000 
      b7       b8       b9      b10 
4.087463 4.169925 4.247928 4.321928

# log2 of each value in each list
rapply(l, log2, how = "list")
$a
 [1] 0.000000 1.000000 1.584963 2.000000 2.321928 2.584963 2.807355 3.000000
 [9] 3.169925 3.321928
$b
 [1] 3.459432 3.584963 3.700440 3.807355 3.906891 4.000000 4.087463 4.169925
 [9] 4.247928 4.321928
 
# what if the function is the mean?
rapply(l, mean)
   a    b 
 5.5 15.5
 
rapply(l, mean, how = "list")
$a
[1] 5.5
$b
[1] 15.5
```

## 7. tapply

# Description: "Apply a function to each cell of a ragged array, that is to each (non-empty) group of values given by a unique combination of the levels of certain factors."

"tapply(X, INDEX, FUN = NULL, ., simplify = TRUE)", where X is "an atomic object, typically a vector" and INDEX is "a list of factors, each of same length as X".

# An ilustrative example. Using iris data, "Species" might be a factor and "iris$Petal.Width" would give us a vector of values. We could then run something like:

```{r}
attach(iris)

# mean petal length by species
tapply(iris$Petal.Length, Species, mean)
    setosa versicolor  virginica 
     1.462      4.260      5.552
```

