---
title: "Managing Data"
author: "Juanjo Moreiras"
date: "Thursday, July 24, 2014"
output: html_document

---

Functions used
rbind                   ## combines rows of data
cbind                   ## combines columns of data
order                   ## sort data frames
merge                   ## match merges two data frames

```
## load Ucla1 to memory

```{r}
load("Ucla1.rda"); str(Ucla1)
```

## Adding Comments

```{r}
## commenting the "Ucla1"
comment(Ucla1)<- "High school and beyond data"

## variable labels using comment
comment(Ucla1$write) <- "writing score"

## Comments stay with our data frame "Ucla1"
save(Ucla1,file="Ucla1.rda")
rm(list=ls())
load(file="Ucla1.rda")
comment(Ucla1); comment(Ucla1$write)
```

## Creating factor variables

```{r}
## Checking if any variables in the Ucla1 are factor
sapply(Ucla1, is.factor)
#   id  female    race     ses  schtyp    prog    read   write    math science   socst 
#  FALSE   FALSE   FALSE   FALSE   FALSE   FALSE   FALSE   FALSE   FALSE   FALSE   FALSE

## Creating factors with value labels
schtypF <- factor(Ucla1$schtyp, levels=c(1,2), labels= c('public', 'private'))
female <- factor(Ucla1$female, levels=c(0,1), labels= c('male', 'female'))
table(schtypF)
#schtypF
# public private 
#    168      32 
table(female)
#female
#  male female 
#    91    109 
```

## Recoding variables  and generating new variables

```{r}
## Recoding race=4 to be NA (missing)
table (Ucla1$race)
# 1   2   3   4 
# 24  11  20 145 

Ucla1$race[Ucla1$race==4] <- NA
table(Ucla1$race)
# 1  2  3 
#24 11 20 

## With missing value
table(Ucla1$race, useNA="ifany")

## Generating a new variable total= read+write
total <- Ucla1$read+ Ucla1$write
summary(total)

## Generating grade variable after recoding total variable
grade <- 0
grade[total <= 75] <- 0
grade[total > 75 & total <= 85] <-1
grade[total > 85 & total <= 90] <-2
grade[total > 90] <- 3

comment(grade) <- "combined grades of read and write"
grade <- factor(grade, levels= c(0,1,2,3),
                labels= c('D','C','B','A'))
table(grade)

## Combining the new varaible total and grade with the original Ucla1 data set
## Using cbind function
Ucla2 <- cbind (Ucla1, cbind(schtypF, female, total, grade))
summary(Ucla2); str(Ucla2)
```

## Keeping and Dropping variables

```{r}
## Keeping only "read" >=60
UclaR <- Ucla1[Ucla1$read >=60,]; UclaR
#     id female race ses schtyp prog read write math science socst
#2   121      1    4   2      1    3   68    59   53      63    61
#4   141      0    4   3      1    3   63    44   47      53    56
#9    84      0    4   2      1    1   63    57   54      58    51
#11   75      0    4   2      1    3   60    46   51      53    61
#13   95      0    4   3      1    2   73    60   71      61    71
#19  114      0    4   3      1    2   68    65   62      55    61
#21  167      0    4   2      1    1   63    49   35      66    41
...........

## Keeping only "read" and "write"
uclaRW <- Ucla1[,c('read', 'write')]; uclaRW
#    read write
#1     57    52
#2     68    59
#3     44    33
#4     63    44
#5     47    52
#6     44    52
#7     50    59
.........

## Another way is with
uclaRW1 <- Ucla1[, c(7,8)]; names(uclaRW1)
# "read"  "write"

## Dropping "read" and "write" using column indices and - signe
uclaDroop <- Ucla1[, -c(7,8)]; names(uclaDroop)
#"id"      "female"  "race"    "ses"     "schtyp"  "prog"    ........ 
```

## Subsetting

```{r}
## Subsetting Ucla1 to two data sets: one female and another male
uclaFemale <- Ucla1[Ucla1$female==1, ]; uclaFemale
#     id female race ses schtyp prog read write math science socst
#2   121      1    4   2      1    3   68    59   53      63    61
#93   82      1    4   3      1    2   68    62   65      69    61
#94    8      1    1   1      1    2   39    44   52      44    48

uclaMale <- Ucla1[Ucla1$female==0, ];uclaMale
#    id female race ses schtyp prog read write math science socst
#1   70      0    4   1      1    1   57    52   41      47    57
#3   86      0    4   3      1    1   44    33   54      58    31
#4  141      0    4   3      1    3   63    44   47      53    56

dim(uclaFemale); dim(uclaMale)

## We put them together again
uclaAll <- rbind(uclaFemale, uclaMale)
dim(uclaAll)
# 200  11
```

### Mergin Files

## We create two data sets from "Ucla1"
## One with demographic variables, and another with other covariates
## We merge both data by the "id" variable

```{r}
uclaDemo <- Ucla1[, c('id','female','race','ses')]
uclaCova <- Ucla1[, c('id','read','write','math','science')]
dim(uclaDemo); dim(uclaCova)

uclaMerge <- merge(uclaDemo, uclaCova, by='id', all=T)
head(uclaMerge)
# id female race ses read write math science
#1  1      1    1   1   34    44   40      39
#2  2      1    1   2   39    41   33      42
#3  3      0    1   1   63    65   48      63
#4  4      1    1   1   44    50   41      39
#5  5      0    1   1   47    40   43      45
#6  6      1    1   1   47    41   46      40

dim(uclaMerge)
```

## If the variable that we were mergin on had different names in each data frame
## Use by.x and by.y arguments
## in by.x list the name of the variable(s) that was in the data frame listed first (uclaDemo)
## in the by.y name variable(s) that was in the second (uclaCova)

```{r}
uclaMerge1 <- merge(uclaDemo, uclaCova, by.x='id', by.y='id', all=T)
uclaMerge1
#     id female race ses read write math science
#1     1      1    1   1   34    44   40      39
#2     2      1    1   2   39    41   33      42
#3     3      0    1   1   63    65   48      63
#4     4      1    1   1   44    50   41      39
#5     5      0    1   1   47    40   43      45
.................
dim(uclaMerge1)
```