---
title: "R-ProgrammCourse"
author: "Juanjo Moreiras"
date: "Tuesday, July 29, 2014"
output: html_document
---
###_______SAMPLE SESSION_________
## Store vectors

```{r}
Height <- c(168, 177, 177, 177, 178, 172, 165, 171, 178, 170)
Height  
#[1] 168 177 177 177 178 172 165 171 178 170

Height[2]         ## Print the second component
#[1] 177

Height[2:5]       ## Print the second, the 3rd, the 4th and 5th component
#[1] 177 177 177 178

(obs <- 1:10)     ## Define a vector as a sequence (1 to 10)
#[1]  1  2  3  4  5  6  7  8  9 10

Weight <- c(88, 72, 85, 52, 71, 69, 61, 61, 51, 75)
BMI <- Weight/((Height/100)^2)   # Performs a simple calculation 
BMI
#[1] 31.17914 22.98190 27.13141 16.59804 22.40879 23.32342 22.40588 20.86112
#[9] 16.09645 25.95156

## Describe the vector with length(), mean() and var()
length(Height)
#[1] 10

mean(Height) ## Compute the sample mean
#[1] 173.3

var(Height)
#[1] 22.23333
```

## Matrix

```{r}
 M <- cbind(obs,Height,Weight,BMI) # Create a matrix
 typeof(M) # Give the type of the matrix
#[1] "double"

 class(M)  ## Give the class of an object
#[1] "matrix"

 is.matrix(M) # Check if   M is a matrix
#[1] TRUE

 is.vector(M)  # M is not a vector
#[1] FALSE

 dim(M)    # Dimensions of a matrix
#[1] 10  4

### For example, to create a  matrix
 matrix(data = NA, nrow = 5, ncol = 5, byrow = T)
#     [,1] [,2] [,3] [,4] [,5]
#[1,]   NA   NA   NA   NA   NA
#[2,]   NA   NA   NA   NA   NA
#[3,]   NA   NA   NA   NA   NA
#[4,]   NA   NA   NA   NA   NA
#[5,]   NA   NA   NA   NA   NA

 matrix(data = 1:15, nrow = 5, ncol = 5, byrow = T)
#     [,1] [,2] [,3] [,4] [,5]
#[1,]    1    2    3    4    5
#[2,]    6    7    8    9   10
#[3,]   11   12   13   14   15
#[4,]    1    2    3    4    5
#[5,]    6    7    8    9   10

## Functions cbind() and rbind() combine vectors into matrices
 v1 <- 1:5
 v2 <- 5:1
 v2
#[1] 5 4 3 2 1

 cbind(v1,v2) ## in a column
#     v1 v2
#[1,]  1  5
#[2,]  2  4
#[3,]  3  3
#[4,]  4  2
#[5,]  5  1

 rbind(v1,v2) ## in a row
#   [,1] [,2] [,3] [,4] [,5]
#v1    1    2    3    4    5
#v2    5    4    3    2    1

## The dimension of a matrix can be obtained using the dim() function.
 X<-matrix(data = 1:15, nrow = 5, ncol = 5, byrow = T)
 dim(X)
#[1] 5 5

 nrow(X) ## number of rows
[1] 5

 ncol(X) ## number of columns
[1] 5

## Unlike data frames matrices must either be numeric or character in type
 a=matrix(2,2,2)
 a
#     [,1] [,2]
#[1,]    2    2
#[2,]    2    2

 a = rbind(a,c("A","A"))
 a
#     [,1] [,2]
#[1,] "2"  "2" 
#[2,] "2"  "2" 
#[3,] "A"  "A"

```

## Arrays
## Composed of n dimensions. Each dimension is a vector of objects of the same type
## An array of one dimension of one element is constructed as

```{r}
## The array x created is single dimension (dim=c(1)) drawn from the vector with values c(T,F).
> x = array(c(T,F),dim=c(1))
x
#[1] TRUE

## A similar array, y, was created with a single dimension and two values.
 y = array(c(T,F),dim=c(2))
 y
#[1]  TRUE FALSE

## A three dimensional array - 3 by 3 by 3 
 z = array(1:27,dim=c(3,3,3))
 dim(z)
#[1] 3 3 3
z
#, , 1
#
#[,1] [,2] [,3]
#[1,]    1    4    7
#[2,]    2    5    8
#[3,]    3    6    9
#
#, , 2
#
#[,1] [,2] [,3]
#[1,]   10   13   16
#[2,]   11   14   17
#[3,]   12   15   18
#
#, , 3
#
#[,1] [,2] [,3]
#[1,]   19   22   25
#[2,]   20   23   26
#[3,]   21   24   27

## Accessing the 3 by 3 by 3 array
## The third dimension is a 3 by 3 array.
 z[,,3]
#[     ,1] [,2] [,3]
#[1,]   19   22   25
#[2,]   20   23   26
#[3,]   21   24   27

### Plot the data using plot()
plot(Height,Weight,ylab="Weight",xlab="Height",main="Corpulence")

```

### data frame

```{r}
 mydat <- data.frame(M)      ## Creates a dataframe
 names(mydat)                ## Give the names of each variable
#[1] "obs"    "Height" "Weight" "BMI"   
 
str(mydat)   
summary(mydat)
View(mydat)                  ## Look at your data

## Save an R session (all the objects in memory) and load the session
 save.image(file="~/Documents/R/test.rda")
 load("~/Documents/R/test.rda")
 dir()                      ## lists content working directory
```

###________SUBSETTING_________

```{r}
## Subsetting Lists
x<- list(foo=1:4,bar=0.6)
x[1]
#$foo
#1234

x[[1]]
#1 2 3 4

x$bar
#0.6
x[["bar"]]
#0.6
x["bar"]
#0.6

## Subsetting multiple elements of a list
x <- list(foo=1:4, bar=0.6,baz="hello")
x[c(1,3)]
#$foo
#[1] 1 2 3 4
#$baz
#[1] "hello"

## Subsetting list with [[ ]]
# We can use just with computed indices
# $ use with literal names
x <- list(foo=1:4,bar=0.6,baz="hello")
# [1] 1 2 3 4

x$name                   ## element "name" doesn't exist
$NULL
x$foo
# [1] 1 2 3 4  # element "foo" does exist

## Subsetting Nested elements of a list
#[[ ]] can take an integer sequence

x <- list(a=list(10,12,14),b=c(3.14,2.81))
x[[c(1,3)]]
# [1] 14   ## same result with
x[[1]][[3]]
# [1] 14

x[[c(2,1)]]
# [1] 3.14

## Partial matching
## Of names with [] and $
x <- list(aardvark=1.5)
x$a
# [1] 1.5

x <-[["a"]]   ##NULL
x[["a",exact=FALSE]]
# [1] 1.5

## Subsetting Missing Values is.na(x)
## Common task is to remove missing values (NAS)
x<- c(1,2,NA,4,NA,5)
bad <- is.na(x) ## I want !NAS
x[!bad]
# [1] 1 2 4 5

## Now I want to take a subset with no missing values
x <- c(1,2,NA,4,NA,5)
y <- c("a","b",NA,"d",NA,"f")
good <- complete.cases(x,y)
good
# [1]  TRUE  TRUE FALSE  TRUE FALSE  TRUE
x[good]
# [1] 1 2 4 5
y[good]
# [1] "a" "b" "d" "f"

## Now for example
## create a vector with 1000 draws from a standard normal distribution
x <- rnorm(1000)

## create a vector with 1000 NAs
z <- rep(NA,1000)

## Now, take a sample of 100 elements at random from these 2000
myData <- sample(c(x,z),100)

## QUESTION: Where our NAs are located in our data?
myNA <- is.na(myData); myNa  ## or with
myData==NA  ## or with ; is.na(myData)

## Counting NA
sum(myNA) ##and finally ; (myData)

## Now, removing NA values from data frames
airquality[1:6,  ]
good <- complete.cases(airquality)
airquality[good, ][1:6, ]  ## with !NA
```
 
###______________CONTROL STRUCTURES__________
## Control structures like: IF, WHILE, ALLOW control the flow of R program
---------
## Using IF (conditional)

```{r}
if(TRUE) {
  print("This is true")
}
else {
  print(This is false")
}
# [1] "This is false"

## This is a valid if/else structure.
if(x > 3) {
        y <- 10
} else {
        y <- 0
}

## So is this one.
y <- if(x > 3) {
        10
} else { 
        0
}

## Of course, the else clause is not necessary.
if(<condition1>) {

}

if(<condition2>) {

}

x <- 2  # x gets the value 2
if(x==3){
    print("This is true")
   }
   else {
     print("This is false")
}
#[1] "This is false"

 if(x==2 && y>2){
     print("x equals 2 and y is greater than 2")
     }
#[1] "x equals 2 and y is greater than 2"
```

## Using LOOPS
## loops with FOR

```{r}
for (i in 1:10) {
   print(i)
}
#[1] 1
#[1] 2
#[1] 3
#[1] 4
#[1] 5
#[1] 6
#[1] 7
#[1] 8
#[1] 9
#[1] 10

x <- c("a", "b", "c", "d")
for(i in 1:4) {  
  print(x[i])
}
#[1] "a"
#[1] "b"
#[1] "c"
#[1] "d"

for(i in seq_along(x)) {
  print(x[i])
}
#[1] "a"
#[1] "b"
#[1] "c"
#[1] "d"

for(letter in x) {
  print(letter)
}
#[1] "a"
#[1] "b"
#[1] "c"
#[1] "d

for(i in 1:4) 
  print(x[i])
#[1] "a"
#[1] "b"
#[1] "c"
#[1] "d

### Nested FOR loops

x <- matrix(1:6, 2, 3)
x
for(i in seq_len(nrow(x))) {
  for(j in seq_len(ncol(x))) {
    print(x[i, j])
  }
}
#[1] 1
#[1] 3
#[1] 5
#[1] 2
#[1] 4
#[1] 6
```

### While loops
## Begin testing a condition, if is true. then execute the loop body. 
## Once loop is executed. Condition is tested again, and so forth

```{r}
 ## With while, sometimes there are multiple conditions
z <- 5
while(z >= 3 && z <= 10) {
  print(z)
  coin <-rbinom(1, 1, 0.5)
  if(coin == 1) {  ## random walk
  z <- z +1
  }else{
    z <- z-1
  }
}
#[1] 5
#[1] 6
#[1] 5
#[1] 4
#[1] 5
#[1] 6
#[1] 7
#[1] 6
#[1] 5
#[1] 4
#[1] 5
#[1] 4
#[1] 5
#[1] 4
#[1] 5
#[1] 6
#[1] 5
#[1] 4
#[1] 5
#[1] 4
#[1] 3

### NEXT in loops
for(i in 1:1000) {
  if(i <= 20) {
    ##Skip the first 20 iterations
    next
  }
  ## Do something here
  }
```

###______vECTORIZATION (implicit loops)__________

### apply() apply a function over the margins of an array
### lapply() Loop over a list and evaluate a function on each element
### sapply() Same as lapply() but try to simplify the resul
### tapply() Apply a function over subsets of a vector
### mapply() multivariate version of lapply()

```{r}
N <- 10
x1 <- rnorm(N)
x2 <- rnorm(N) + x1 + 1
male <- rbinom(N,1,.48)
y <- 1 + xi + x2 + male + rnorm(N)
myData <- data.frame(y,x1,x2,male)
lapply(myData,mean) #returns a list
#$y
#[1] 4
#$x1
#[1] 0.2770778
#$x2
#[1] 1.018787
#$male
#[1] 0.3

sapply(myData,mean) #returns a vector
#y        x1        x2      male 
#4.0000000 0.2770778 1.0187873 0.3000000 

apply(myData,1,mean) # applies the function to each row
#[1]  1.1171547  0.8343634 -0.3248893  1.6203620  1.4766708  2.1996836
#[7]  2.2057034  2.0249858  1.2975479  1.5380803
apply(myData,2,mean) # applies the function to each column
#4.0000000 0.2770778 1.0187873 0.3000000 

tapply(myData$y,myData$male,mean) #apllies the function to each level of a factor
#    0     1 
#1.040 5.454

### lapply() always will return a list
x <- list(a = 1:5,b = rnorm(10))
lapply(x,mean)
# [1] 3
#
#$b
#[1] 0.1570857

x <- list(a = 1:4,b = rnorm(10), c = rnorm(20,1), d=rnorm(100,5))
lapply(x,mean)
#$a
#[1] 2.5
#$b
#$[1] 0.140962
#$c
#[1] 0.8963297
#$d
#[1] 5.153206

x <- 1:4
lapply(x,runif, min = 0, max = 5)
#[[1]]
#[1] 2.090001
#
#[[2]]
#[1] 2.582228 4.109156
#
#[[3]]
#[1] 3.521435 2.639731 1.516384
#
#[[4]]
#[1] 0.5536162 4.6146353 2.1220014 4.9074133

## saplly() will try to simplify the result of lapply() if possible
x <- list(a = 1:4,b = rnorm(10), c = rnorm(20,1), d=rnorm(100,5))
sapply(x,mean)
#        a         b         c         d 
#2.5000000 0.2157379 0.7183672 5.0590031

## apply() is most often use to apply a function to rows or columns of a matrix
## it can be used with general arrays
x <- matrix(rnorm(200),20,10)
apply(x,2,mean)
#[1] -0.36878457  0.14632712  0.29989357 -0.03243227 -0.34132280  0.09424873
#[7]  0.41093092 -0.02263104 -0.20314875  0.33647587

apply(x,1,sum)
#[1]  -0.162338828  2.873914227 -1.998521116 -0.675386191  6.149962253
#[6]   1.764600237  1.914994782 -0.243225568  1.323827626 -2.008820033
#[11] -0.297638953  5.245163684  1.387333135 -1.563486948 -4.286454776
#[16]  0.890275692 -1.324523170  0.009703546  0.853276663 -3.461520357

### apply() col/wow sums and means
## rowSums  = apply(x,1,sum)
## rowMeans = apply(x,1,mean)
## colSums  = apply(x,2,sum)
## colMeans = apply(x,2,mean)

### Apply() other ways. 
### Quantiles of rows of a matrix
x <- matrix(rnorm(200),20,10)
apply(x,1,quantile,probs = c(0.25,0.75))
#          [,1]       [,2]       [,3]        [,4]       [,5]      [,6]
#25% -1.1382083 -0.7154221 -0.3450960 -0.93566369 -0.1546529 0.1450603
#75%  0.8532476  0.2737537  0.8920855  0.06372081  1.5154800 0.7822728
#          [,7]       [,8]       [,9]      [,10]      [,11]      [,12]
#25% -0.7015191 -0.4276700 -0.3998783 -0.9788060 -0.1306109 -0.5242083
#75%  0.7835900  0.9008524  0.6311825 -0.6679648  1.2437027  0.8375514
#         [,13]      [,14]      [,15]      [,16]      [,17]      [,18]
#25% -0.2704254 0.03811542 -0.3599430 -0.4430957 -0.3879781 -0.7469107
#75%  0.8724733 0.59088888  0.4159396  0.3341540  0.7705576  0.2148826
         #[,19]      [,20]
#25% -0.2671408 -0.4518031
#75%  0.6407976  0.3521868

### Average matrix in an array
a <- array(rnorm(2 * 2 * 10), c(2,2,10))
apply(a,c(1,2),mean)
#           [,1]         [,2]
#[1,]  0.6369863 -0.008836067
#[2,] -0.1583577  0.679116073
rowMeans(a,dims = 2)
#           [,1]         [,2]
#[1,]  0.6369863 -0.008836067
#[2,] -0.1583577  0.679116073

##tapply() is used to apply a function over subsets of a vector
x <- c(rnorm(10),runif(10),rnorm(10,1))
f <- gl(3,10)
f
#[1] 1 1 1 1 1 1 1 1 1 1 2 2 2 2 2 2 2 2 2 2 3 3 3 3 3 3 3 3 3 3
#Levels: 1 2 3
tapply(x,f,mean)
#         1          2          3 
#-0.2159529  0.5633908  0.6797613 

## tapply() Take group means without simplification
tapply(x,f,mean,simplify = FALSE)
#$`1`
#[1] -0.2159529
#
#$`2`
#[1] 0.5633908
#
#$`3`
#[1] 0.6797613

## tapply() find group ranges
tapply(x,f,range)
#$`1`
#[1] -1.813917  1.367771
#
#$`2`
#[1] 0.09509871 0.92299818
#
#$`3`
#[1] -0.8293094  1.7328687

## mapply multivariable apply of sorts
## the following is tedious to type
list(rep(1,4),rep(2,3),rep(3,2),rep(3,1))
## instead we can do
mapply(rep,1:4,4:1)
#[[1]]
#[1] 1 1 1 1
#
#[[2]]
#[1] 2 2 2
#
#[[3]]
#[1] 3 3
#
#[[4]]
#[1] 4

## mapply() Instant vectorization
noise <- function(n,mean,sd) {
  rnorm(n,mean,sd)
}
noise(5,1,2)
}
noise(5,1,2)
[1]  2.8199031  1.4544538 -1.1531074  0.8360228  4.0756294

noise(1:5,1:5,2)
[1] 3.517232 3.878232 3.911837 3.688214 9.424089
## Instant vectorizacion with mapply()
mapply(noise, 1:5,1:5,2)
#[[1]]
#[1] 0.6724438
#
#[[2]]
#[1] 2.674333 5.004172
#
#[[3]]
#[1] -0.5019022  0.2058145  3.6292665
#
#[[4]]
#[1]  5.395888  2.151710  3.524492 -1.551377
#
#[[5]]
#[1] 5.437492 4.195407 9.070071 5.043088 3.974678

## Which is the same instant vectorization as
list(noise(1,1,2),noise(2,2,2),
     noise(3,3,2),noise(4,4,2),
     noise(5,5,2))

### split() takes a vector or other objects and splits it into groups
## determined by a factor or list of factors
x <- c(rnorm(10), runif(10),rnorm(10,11))
f <- gl(3,10)
split(x,f)
#$`1`
#[1]  1.09966243 -0.86458790 -0.46436673  0.74972632 -0.02083017  0.41624768
#[7] -2.62977136 -1.66529941 -0.71587141 -0.19667895
#
#$`2`
#[1] 0.5182775 0.4610488 0.8149972 0.0552355 0.9279215 0.4860266 0.5653717
#[8] 0.5515969 0.5982286 0.4903615
#
#$`3`
#[1]  9.665972 12.863321 12.517250 12.399225 12.667451 11.899372 11.886343
#[8]  9.794893 12.107294  9.761322

## A common idiom is split() followed by an lapply()
lapply(split(x,f),mean)
#$`1`
#[1] -0.429177
#
#$`2`
#[1] 0.5469066
#
#$`3`
#[1] 11.55624

## Splitting a data frame
library(datasets)
head(airquality)
#Ozone Solar.R Wind Temp Month Day
#1    41     190  7.4   67     5   1
#2    36     118  8.0   72     5   2
#3    12     149 12.6   74     5   3
#4    18     313 11.5   62     5   4
#5    NA      NA 14.3   56     5   5
#6    28      NA 14.9   66     5   6
 s <- split(airquality,airquality$Month)
lapply(s, function(x) colMeans(x[,c("Ozone","Solar.R","Wind")]))
#$`5`
#  Ozone  Solar.R     Wind 
#     NA       NA 11.62258 
#
#$`6`
#  Ozone   Solar.R      Wind 
#     NA 190.16667  10.26667 
#
#$`7`
#  Ozone    Solar.R       Wind 
#     NA 216.483871   8.941935 
#
#$`8`
#  Ozone  Solar.R     Wind 
#     NA       NA 8.793548 
#
#$`9`
#  Ozone  Solar.R     Wind 
#     NA 167.4333  10.1800 

### Splitting a data frame
sapply(s,function(x)colMeans(x[,c("Ozone","Solar.R","Wind")]))
#               5         6          7        8        9
#Ozone         NA        NA         NA       NA       NA
#Solar.R       NA 190.16667 216.483871       NA 167.4333
#Wind    11.62258  10.26667   8.941935 8.793548  10.1800

sapply(s,function(x)colMeans(x[,c("Ozone","Solar.R","Wind")],
                             na.rm = TRUE))
#Ozone    23.61538  29.44444  59.115385  59.961538  31.44828
#Solar.R 181.29630 190.16667 216.483871 171.857143 167.43333
#Wind     11.62258  10.26667   8.941935   8.793548  10.18000

### Splitting on more than one level
x <- rnorm(10)
f1 <- gl(2,5)
f2 <- gl(5,2)
f1
#[1] 1 1 1 1 1 2 2 2 2 2
#Levels: 1 2
f2
#[1] 1 1 2 2 3 3 4 4 5 5
#Levels: 1 2 3 4 5
interaction(f1,f2)
#[1] 1.1 1.1 1.2 1.2 1.3 2.3 2.4 2.4 2.5 2.5
#Levels: 1.1 2.1 1.2 2.2 1.3 2.3 1.4 2.4 1.5 2.5

## Interaction can create empty levels
str(split(x,list(f1,f2)))
#List of 10
#$ 1.1: num [1:2] 1.104 0.769
#$ 2.1: num(0) 
#$ 1.2: num [1:2] -0.649 0.469
#$ 2.2: num(0) 
#$ 1.3: num -0.246
#$ 2.3: num 0.766
#$ 1.4: num(0) 
#$ 2.4: num [1:2] 1.3 -0.453
#$ 1.5: num(0) 
#$ 2.5: num [1:2] -0.194 0.104

## empty levels can be dropped
str(split(x,list(f1,f2),drop=TRUE))
#$ 1.1: num [1:2] 1.104 0.769
#$ 1.2: num [1:2] -0.649 0.469
#$ 1.3: num -0.246
#$ 2.3: num 0.766
#$ 2.4: num [1:2] 1.3 -0.453
#$ 2.5: num [1:2] -0.194 0.104
```

###__________DEBUGGING TOOLS______________

## traceback: prints out the function cal stack. does nothing if no error
## debug: flags a function
## browser: suspends the execution of a function and puts in debug mode
## trace: allows to insert debugging code into a function
## recoves: allows to modify the error

## Tracing errors with traceback()
## Example the following function whose sole purpose is to throw an error.

```{r}
myFun <- function () {
  stop("Woops! an error")
}

## A call to myFun() gives
myFun()
#Error in myFun() : Woops! an error

## Calling myFun2() and traceback() gives
 myFun2()
#Error in myFun() : Woops! An error
 traceback()
#3: stop("Woops! An error")
#2: myFun()
#1: myFun2()

## Another example
 mean(m)
# Error in mean(m) : object 'm' not found
traceback()
# 1: mean(m)
```

###__________SIMULATION______________

### rnorm: generate random Normal variates with a given mean and sd
##  rnorm(n, mean = 0, sd = 1)
### pnorm: evaluate the cumulative distribution function for a Normal distribution
##  pnorm(q,mean = 0,sd = 1,lower.tail = TRUE, log.p = FALSE)
### dnorm: evaluate the Normal probability density at a point or vector of points
##  dnorm(x,mean = 0, sd = 1. log = FALSE)
### rpois: generate random Poisson variates with a given rate

```{r}
## Generating random numbers
x <- rnorm(10)
x
#[1]  1.39428733 -0.34308042  0.12116540  0.09302132 -1.20947813 -0.15097663
#[7] -0.09176090  1.16338136 -2.33987778  2.47667709

set.seed(1)
x <- rnorm(10,20,2)
x
#[1] 24.64946 18.64844 23.55926 23.32695 21.61096 19.51958 19.31132 19.43904
#[9] 16.40549 20.99029
summary(x)
#Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#16.41   19.34   20.25   20.75   22.90   24.65 

## Generating Poisson data
rpois (10, 1)
#[1] 2 1 1 0 3 1 0 0 0 1
rpois(10,2)
#[1] 0 1 3 3 0 0 1 1 1 5
rpois(10,20)
#[1] 23 17 16 20 16 17 14 21 23 24

ppois(2,2) ##cumulative distribution
#[1] 0.6766764 ##Pr(x <=2)
ppois(4,2)
#[1] 0.947347  ##Pr(x <= 4)
ppois(6,2)
#[1] 0.9954662  ##Pr(x <= 6)

## Generating random numbers from linear model
set.seed(20)
x <- rnorm(100)
e <- rnorm(100,0,2)
y <- 0.5 + 2 * x + e
summary(y)
#    Min. 1st Qu.  Median    Mean  3rd Qu.    Max. 
#-6.4080 -1.5400   0.6789  0.6893  2.9300  6.5050
plot(x,y)

## Generating random numbers from linear model
## x is binary
set.seed(10)
x <- rbinom(100,1,0.5)
e <- rnorm(100,0,2)
y <- 0.5 +2 *x + e
summary(y)
#    Min. 1st Qu. Median    Mean  3rd Qu.    Max. 
#-3.4940 -0.1409  1.5770  1.4320  2.8400  6.9410 
plot(x,y)


## Generating random numbers from a generalized linear model
## Simulating a Poisson model
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#0.00    1.00    1.00    1.55    2.00    6.00 
plot(x,y)

## Random sampling from arbitrary distributions
set.seed(1)
sample(1:10,4)
#[1] 7 1 4 8
sample(letters,5)
#[1] "j" "y" "e" "s" "b"
sample(1:10)   ##permutation
#[1]  5  2  9  6 10  8  7  1  3  4
sample(1:10)
#[1]  7  6  3  8  9  5  4 10  2  1
sample(1:10,replace = TRUE)  ## sample w/replacement
#[1]  8 10  3  4  9  1  4  5  2  6
```

###______________________TIMES AND DATES____________________________

### The default format is yyyy-mm-dd hh:mm:ss or %Y-%m-%d %H:%M:%S
### For example 2014-07-12 23:12:24

```{r}
### System date and time
Sys.time()
#[1] "2014-07-12 15:41:09 EDT"

format(Sys.time(),"%H %M")   # in a different format and without the date
#[1] "15 41"

Sys.Date()
#[1] "2014-07-12"
 date()                       # returns the current date and time,
#[1] "Sat Jul 12 15:42:46 2014"

## Convert strings to date/time objects
myDate <- as.Date("2014-07-12")
myDate
#[1] "2014-07-12"

## input date in a different format
myDate2 <- as.Date("2014/07/12", format="%m/%d/%y") 
myDate2

## input time in a different format
myTime <- strptime("14/07/12 14.34.35", format="%m/%d/%y %H.%M.%S")
myTime

## convert a date/time object to a normal string
myString <- as.character(Sys.time())
myString

## Extracting information from dates
## Get a string representing the weekday of the specified date
 weekdays(myDate)
#[1] "Saturday"
 months(my.date)
#[1] "July"
myDate
#[1] "2014-07-12"
## the integer number of days since the beginning of epoch
julian(my.date)
#[1] 16263
#attr(,"origin")
#[1] "1970-01-01"

```
## Dates in R
## Dates are represented by the Date class
## Can be coerced from a character string using the as.Date() function.

```{r}
x <- as.Date("1970-01-01")
x
#[1] "1970-01-01"

### Times in R
## Times are represented using the POSIXct or the POSIXlt class
## Can be coerced from a character string using as.POSIXlt or as.POSIXct function.
x <- Sys.time()
x
#[1] "2014-07-12 21:55:29 EDT"
p <- as.POSIXlt(x)
names(unclass(p))
#[1] "sec"    "min"    "hour"   "mday"   "mon"    "year"   "wday"   "yday"   "isdst" 
#[10] "zone"   "gmtoff"
p$sec
#[1] 29.34113

## You can also use the POSIXct format.
x <- Sys.time()
x    ## Already in 'POSIXct' format
#[1] "2014-07-12 22:04:17 EDT"
unclass(x)
#[1] 1405217057
#Error: $ operator is invalid for atomic vectors
p <- as.POSIXlt(x)
p$sec
#[1] 17.18532

## There is strptime function function in case dates are written in different format
 datestring <- c("January 10, 2012 10:40")
 x <- strptime(datestring, "%B %d, %Y %H:%M")
 x
#[1] "2012-01-10 10:40:00 EST"
class(x)
#[1] "POSIXlt" "POSIXt" 

## Operations on Dates and Times
##You can use mathematical operations on dates and times
x <- as.Date("2012-01-01")
y <- strptime("9 Jan 2011 11:34:21", "%d %b %Y %H:%M:%S") 
x <- as.POSIXlt(x) 
x-y
#Time difference of 356.3095 days

## Even keeps track of leap years, leap seconds, daylight savings, and time zones.
x <- as.Date("2012-03-01") ; y <- as.Date("2012-02-28") 
x-y
#Time difference of 2 days

x <- as.POSIXct("2012-10-25 01:00:00")
y <- as.POSIXct("2012-10-25 06:00:00", tz = "GMT") 
y-x
#Time difference of 1 hours