---
title: "tapply and split"
author: "Juanjo Moreiras"
date: "Friday, July 25, 2014"
output: html_document


## tapply (Roger Peng). To apply a function over subsets of a vector

# function (X, INDEX, FUN = NULL, ..., simplify = TRUE)
- `X` is a vector
- `INDEX` is a factor or a list of factors (or else they are coerced to factors) 
- `FUN` is a function to be applied
- ... contains other arguments to be passed `FUN`
- `simplify`, should we simplify the result?

```{r Roger}

> str(tapply)
#function (X, INDEX, FUN = NULL, ..., simplify = TRUE)

```

## Take group means.

```{r Roger}

x <- c(rnorm(10), runif(10), rnorm(10, 1)); x
# [1] -2.27265022  1.96454972  0.19220818  0.87602574  0.71518988 -0.22076331  0.45142154
# [8]  0.06268169  1.05329294  2.04540557  0.88660834  0.87920589  0.86578012  0.29967751
#[15]  0.96081425  0.63754863  0.43947666  0.03630077  0.61804896  0.45872447 -0.05390791
#[22]  2.31971734  1.04949095 -0.03544434  1.21714768  2.02673117  1.46553506  0.50373965
#[29]  1.51163628 -1.59413027

f <- gl(3, 10); f 
# [1] 1 1 1 1 1 1 1 1 1 1 2 2 2 2 2 2 2 2 2 2 3 3 3
#[24] 3 3 3 3 3 3 3
#Levels: 1 2 3

tapply(x, f, mean)
        1         2         3 
0.1144464 0.5163468 1.2463678

```

## Take group means without simplification.

```{r Roger}

tapply(x, f, mean, simplify = FALSE)
#$`1`
#[1] 0.4867362

#$`2`
#[1] 0.6082186

#$`3`
#[1] 0.8410516

```

## Find group ranges.

```{r Roger}

tapply(x, f, range)
#$`1`
#[1] -2.272650  2.045406

#$`2`
#[1] 0.03630077 0.96081425

#$`3`
#[1] -1.594130  2.319717

```

## split. takes a vector or other objects and splits it into groups determined by a factor or list of factors

# function (x, f, drop = FALSE, ...)

- `x` is a vector (or list) or data frame
- `f` is a factor (or coerced to one) or a list of factors
- `drop` indicates whether empty factors levels should be dropped

```{r Roger}

str(split)
function (x, f, drop = FALSE, ...)
  
```  

## An a example of split

```{r Roger}

x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
split(x, f)
#$`1`
# [1]  0.04369846  0.64613515  1.34142474 -0.74953436 -1.13573961 -1.22188769  0.86610468  0.99275201  3.23702995
# [10]  0.64253753
#$`2`
# [1] 0.6116775 0.6327786 0.8004982 0.5051107 0.4092149 0.9231836 0.5768224 0.8965029 0.1298950 0.9934782
#$`3`
# [1]  2.0673728  0.9573406  2.2214485  1.2911139  1.1308204 -0.6184346  1.1317584  1.7641603  2.3844945  0.9834543

```

## A common idiom is `split` followed by an `lapply`.

```{r Roger}

lapply(split(x, f), mean)
#$`1`
#[1] -0.2300855
#$`2`
#[1] 0.4801524
#$`3`
#[1] 0.7864772

```

## Splitting a Data Frame named 'airquality'

```{r Roger}

library(datasets)
head(airquality)
#  Ozone Solar.R Wind Temp Month Day 
#1    41     190  7.4   67     5   1 
#2    36     118  8.0   72     5   2 
#3    12     149 12.6   74     5   3 
#4    18     313 11.5   62     5   4 
#5    NA      NA 14.3   56     5   5 
#6    28      NA 14.9   66     5   6

## With split and lapply a data frame
s <- split(airquality, airquality$Month)
lapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))
#$`5`
#   Ozone  Solar.R     Wind 
#      NA       NA 11.62258 
#$`6`
#    Ozone   Solar.R      Wind 
#       NA 190.16667  10.26667 
#$`7`
#     Ozone    Solar.R       Wind 
#        NA 216.483871   8.941935 
#$`8`
#   Ozone  Solar.R     Wind 
#      NA       NA 8.793548 
#$`9`
#   Ozone  Solar.R     Wind 
#      NA 167.4333  10.1800 

## with sapply
sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")])) 
#               5         6          7        8        9
#Ozone         NA        NA         NA       NA       NA
#Solar.R       NA 190.16667 216.483871       NA 167.4333
#Wind    11.62258  10.26667   8.941935 8.793548  10.1800

## with sapply and na.rm=T
sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")],
                                 na.rm = TRUE))
                  5            6             7            8           9 
#Ozone      23.61538     29.44444     59.115385    59.961538   31.44828 
#Solar.R   181.29630    190.16667    216.483871   171.857143  167.43333 
#Wind       11.62258     10.26667      8.941935     8.793548   10.18000

```

## Splitting on more than One Level

```{r Roger}

x <- rnorm(10); x
#[1]  0.62642194  0.04022217 -0.34424828 -0.18752353 -0.71192691 -0.03807643  0.20515955 -1.00306761  0.77065719 -0.48262475
f1 <- gl(2, 5)
f2 <- gl(5, 2)
f1
# [1] 1 1 1 1 1 2 2 2 2 2
#Levels: 1 2
f2
# [1] 1 1 2 2 3 3 4 4 5 5
#Levels: 1 2 3 4 5
interaction(f1, f2)
# [1] 1.1 1.1 1.2 1.2 1.3 2.3 2.4 2.4 2.5 2.5
#10 Levels: 1.1 2.1 1.2 2.2 1.3 2.3 1.4 ... 2.5

## Interactions can create empty levels
str(split(x, list(f1, f2)))
#List of 10
# $ 1.1: num [1:2] -0.378  0.445
# $ 2.1: num(0)
# $ 1.2: num [1:2] 1.4066 0.0166
# $ 2.2: num(0)
# $ 1.3: num -0.355
# $ 2.3: num 0.315
# $ 1.4: num(0)
# $ 2.4: num [1:2] -0.907  0.723
# $ 1.5: num(0)
# $ 2.5: num [1:2] 0.732 0.360


